---
title: "Once Upon a Stormy Evening"
execute:
  warning: false
---

# Introduction

The purpose of this HW is for you to utilize your descriptive statistics skills to a real world dataset and problem. Here we will investigate

1\) how the spatial distribution and frequency of different severe weather events in South Carolina has changed over time, and also

2\) explore the vulnerability of communities to these different severe weather events.

Luckily for us, the National Center for Environmental Information at NOAA maintains a comprehensive database of severe weather events. You can read more about it here: <https://www.ncei.noaa.gov/stormevents/>

## Load packages and data

```{r}
library(tidyverse)
library(sf)
library(tigris)
options(tigris_use_cache = TRUE)
library(plotly)

# load SC sever weather events 2015 and 2024

dat15 <- read_csv("data/sc_nhaz_15.csv")
dat24 <- read_csv("data/sc_nhaz_24.csv")
```

This is a cleaned up and truncated version of the actual dataset which is very messy. However, even though this data is 'clean', there can be several hidden issues. To know what these variables are, refer to the codebook (Page 1-6) in the docs folder.

Q.1 Your first main research question that you are trying to answer is about [how the spatial distribution and frequency of different severe weather events in South Carolina has changed over time]{.underline}.

[**Based on the data and variables provided to you,**]{.underline}what could be 3 descriptive questions about these severe weather events that you could ask related to your main research question. Atleast one of these questions should be spatial in nature.

For example, one potential non-spatial descriptive question is: What is the difference in average cost of property damage caused by floods between 2015 and 2024? A potential spatial descriptive question could be: What is the spatial distribution of flooding events compared to extreme heat events in 2024?

Q.2 Examine the following variables more closely. Explain what kind of variable they [should be,]{.underline} and how [they are actually stored a]{.underline}s in the dataset.

1.  EVENT_TYPE

2.  DEATHS_DIRECT

3.  DAMAGE_PROPERTY

Q.3 Now we will make some modifications to the dataset.

```{r}
# create a new column for property damage to store values in USD
dat15 <- dat15 |>
  mutate(PROP_DAMAGE_USD = parse_number(DAMAGE_PROPERTY)*1000,
         ALL_INJ = INJURIES_DIRECT + INJURIES_INDIRECT,
         ALL_DEATH = DEATHS_DIRECT + DEATHS_INDIRECT)

## your code here to do the same for 2024
dat24 <- dat24 |>
  mutate(PROP_DAMAGE_USD = parse_number(DAMAGE_PROPERTY)*1000,
         ALL_INJ = INJURIES_DIRECT + INJURIES_INDIRECT,
         ALL_DEATH = DEATHS_DIRECT + DEATHS_INDIRECT)
```

Q.4 Lets answer a few descriptive questions:

1.  What were the top 3 types of severe weather events in 2015? What were the top 3 types of severe weather events in 2024? Were there any differences?

2.  Which event had the highest property damage in 2015 and 2024 in SC? Which event had the highest injuries and deaths in 2015 and 2024?

3.  What was the mean/average property damage categorized by event type in 2015 and 2024? What was the median property damage categorized by event type in 2015 and 2024? What was the standard deviation for the variation in property damage?

4.  Based on the previous answer, describe one event type for which the average property damage increases, and one for which it decreased.

Q.5 Now convert the csv files to spatial data

```{r}
dat15 <- st_as_sf(dat15, coords = c("BEGIN_LON","BEGIN_LAT"), crs = 4269)
dat24 <- st_as_sf(dat24, coords = c("BEGIN_LON","BEGIN_LAT"), crs = 4269)

## Import SC county file
county_geo <- counties(state = "SC")
```


```{r}
map15 <- ggplot() + # tell R that you are creating a graphic
  geom_sf(data = county_geo) +
  geom_sf(data = dat15,
          color = "red", # assign red color to each point
          alpha = 0.2) +
  geom_sf_label(data = county_geo,aes(label = NAME), size = 2) +
  theme_minimal()# tell R what GIS layer you are using.
plot(map15)
```


```{r}
source("code/descriptive_geo_functions.R")

dat15_deaths <- calc_mc(deaths_loc)
```

