---
title: "Homework 1: Once Upon a Stormy Evening"
author: "Varun Goel"
description: |
  Due Date: **Sunday, October 19, 5 pm** \newline \newline
  **(250 points + 20 bonus points)**
execute:
  warning: false
  eval: false
date: "`r Sys.Date()`"
format: 
  html:
    toc: true
    toc-expand: 3
editor: visual
---

# Introduction

The purpose of this HW is for you to utilize your descriptive statistics skills to a real world dataset and problem. Here we will investigate how the spatial distribution and frequency of different severe weather events in South Carolina has changed over time.You will also spend time on your final project question formulation and dataset finding.

# Instructions

-   Go to `File –> Save as`, and save this as `hw1_answers.qmd`. Then you can use that file for filling in all your answers.

-   **Very important:**

    -   In the header of the `.qmd` file, change name under `author:` to your name.

    -   change `eval: false` to `eval: true`

-   Please make sure that your code is entered in your code block but your written answers are OUTSIDE your codeblock. Only put comments (using #) in your code to describe your code.

-   [**Read carefully:**]{.underline} You are welcome to use ChatGPT or other generative AI for your assignment. I encourage to use it mainly for coding help as needed. However, I also encourage you to first try code yourself based on what we have learned so that you can get some practice. If you do use generative AI [**for any part of the assignment,**]{.underline}create a word doc titled `ai_prompts.docx`. For [**each prompt**]{.underline} that you provide to generative AI software, report the following in the document:

    -   The Question number from your homework

    -   Your exact prompt as inputted to the AI

    -   The [RAW (]{.underline}unchanged) answer generated by the software)

-   If you do use AI, but do not disclose [**ALL**]{.underline} your prompts , [**I will consider it as plagiarism and take strict action including giving you a zero on your assignment**]{.underline}. There is a specific manner in which softwares like ChatGPT write code. If I detect those patterns or suspect you used AI but did not report it, I may ask you to solve the problem in front of me to ensure that you did the problem and not generative AI. Hence, to be safe, please report all your prompts even if you made little or extensive use of generative AI.

-   [You are welcome to work with a classmate or discuss answers and code, but your homework should be your own work]{.underline}. If you do work with a classmate, please report in the document whom you worked with so that I give you the benefit of doubt in case there are similar answers with an another person's work.

-   DONT WAIT till the Last Minute. Ask me for help, proactively. I am here for you.

## 1 Load packages and data

Luckily for us, the National Center for Environmental Information at NOAA maintains a comprehensive database of severe weather events. You can read more about it here: <https://www.ncei.noaa.gov/stormevents/>

```{r}
#load packages
library(tidyverse)
library(sf)
library(plotly)
library(tidycensus) # for reading census data in R
library(tigris)# For reading geographic admin boundaries in R
options(tigris_use_cache = TRUE)

### DONT TOUCH THIS NEXT LINE OF CODE ###
census_api_key(key = "a0b70fd4d6380a164f8b90cef1e114f9a7ed6706", install = TRUE, overwrite = TRUE)
# load SC severe weather events 2015 and 2024
dat15 <- read_csv("data/sc_nhaz_15.csv")
dat24 <- read_csv("data/sc_nhaz_24.csv")
```

This is a cleaned up and truncated version of the actual dataset which is very messy. However, even though this data is 'clean', there can be several hidden issues. To know what these variables are, refer to the [**codebook (Page 1-6) in the docs folder**]{.underline}.

[Q.1 Your main research question that you are trying to answer is about [how the spatial distribution and frequency of different severe weather events in South Carolina has changed over time]{.underline}.]{style="color: darkred;"}

[[**Based on the data and variables provided to you,**]{.underline} what could be 3 descriptive questions about these severe weather events that you could ask related to your main research question. Atleast one of these questions should be spatial in nature.]{style="color: darkred;"}**\[5 points\]**

For example, one potential non-spatial descriptive question is: What is the difference in average cost of property damage caused by floods between 2015 and 2024? A potential spatial descriptive question could be: What is the spatial distribution of flooding events compared to extreme heat events in 2024?

## 2 Examining your Data

### 2.1 Cleaning Data

[Q.2 Examine the following variables more closely. Explain what kind of variable (nominal, ordinal, interval, ratio) they [should be,]{.underline} and how [they are actually stored a]{.underline}s in the dataset.]{style="color: darkred;"}**\[5 points\]**

1.  [EVENT_TYPE]{style="color: darkred;"}

2.  [DEATHS_DIRECT]{style="color: darkred;"}

3.  [DAMAGE_PROPERTY]{style="color: darkred;"}

[Q.3 Now we will make some modifications to the dataset. Answer the following coding question for the 2024 data set:]{style="color: darkred;"} **\[5 points\]**

```{r}

# We will first create new variables or fix old variables
# create a new column that stores property damage values in USD
# also create a column that combines injuries, and deaths
# convert event types to factor class

dat15 <- dat15 |>
  mutate(PROP_DAMAGE_USD = parse_number(DAMAGE_PROPERTY)*1000,
         ALL_INJ = INJURIES_DIRECT + INJURIES_INDIRECT,
         ALL_DEATH = DEATHS_DIRECT + DEATHS_INDIRECT,
         EVENT_TYPE_F = as.factor(EVENT_TYPE))

## RUN YOUR CODE HERE to do the same for dat24
dat24 <-
```

Now, for this exercise we will only keep events which do not have missing property damage data.

```{r}
dat15 <- dat15 |> filter(!is.na(PROP_DAMAGE_USD))
dat24 <- dat24 |> filter(!is.na(PROP_DAMAGE_USD))
```

### 2.2 Descriptive Statistics

[Q.4 Lets answer a few descriptive questions:]{style="color: darkred;"}

1.  [What were the top 3 types of severe weather events overall?]{style="color: darkred;"} [What were the top 3 types of severe weather events in 2015? What were the top 3 types of severe weather events in 2024? Were there any differences?]{style="color: darkred;"} **\[15 points\]**

    ```{r}
    # your code here
    ```

2.  [Which event had the highest property damage in 2015 and 2024 in SC? Which event had the highest injuries and deaths in 2015 and 2024?]{style="color: darkred;"} **\[10 points\]**

    ```{r}
    # your code here
    ```

3.  [What was the mean/average property damage categorized by event type in 2015 and 2024? What was the median property damage categorized by event type in 2015 and 2024? What was the standard deviation for the variation in property damage?]{style="color: darkred;"} **\[15 points\]**

    ```{r}
    # your code here
    # for standard deviation, you can use the sd() function.
    ```

4.  [Based on the previous answer, describe one event type for which the average property damage increases, and one for which it decreased.]{style="color: darkred;"} **\[10 points\]**

    ```{r}
    # your code here
    ```

### 2.3 Examining Spatial Data

Since we have locational information in our dataframe, we will convert the dataframe into a spatial object.

Q.5 [Now convert the csv files to spatial data]{style="color: darkred;"} **\[5 points\]**

```{r}
dat15 <- st_as_sf(dat15, coords = c("BEGIN_LON","BEGIN_LAT"), crs = 4269) %>%
  st_transform(crs = 32133) # Here we convert Geographic projection to a Lambert Conformal Conic Projection. This helps use deal with area level data better

## YOUR CODE HERE for dat24
dat24 <- 

```

### 2.4 Retrieving Spatial Data

One great part about coding and the R ecosystem is that you can directly retrieve some spatial data without going through the hassle of downloading and storing it. It is beyond the scope of this class, but the census bureau has APIs (Application Programming Interface) that let you 'connect' to their data ecosystem and get all sorts of tabular and spatial data. The `tigris` and the `tidycensus` packages help read that data in R. Here we will retrieve SC county level geospatial data using that functionality. We can then use this data on our map to provide more context.

```{r}
## Import SC county file from the census api
county_geo <- counties(state = "SC") %>%
  st_transform(crs = 32133)
```

### 2.5 Mapping Spatial Distribution of extreme events

Our first step towards spatially exploring our data is to map the distribution of extreme events. Using `ggplot()` we can make various kinds of maps. Here is a simple map

```{r}
map15_1 <- ggplot() + # tell R that you are creating a graphic
  geom_sf(data = county_geo) + # create a base layer of counties
  geom_sf(data = dat15,# add the extreme event points
          color = "red", # assign red color to each point
          alpha = 0.2) + # for transparency
  geom_sf_label(data = county_geo,aes(label = NAME), size = 2) + # add labels to counties
  theme_minimal() +
  coord_sf(datum = st_crs(county_geo)) # ensure map is in right coordinate system
plot(map15_1)
```

In the first map, we assume that each extreme event is exactly the same. Is that a realistic assumption. In the next map, lets create a proportional symbol map where we the size of dots based on how much property damage an event did. We can also use colors to highlight events with more property damage vs less. There are lots of colors one can use, but nowadays cartographers are using color schemes called 'viridis' that are universally accepted.

<https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html>

```{r}
map15_2 <- ggplot() + 
  geom_sf(data = county_geo) +
  geom_sf(data = dat15,
          aes(size = PROP_DAMAGE_USD, color = PROP_DAMAGE_USD),# size and color are based on property damage
          alpha = 0.2) +
  scale_color_viridis_c(option = "magma",
                        direction = -1) + # there are several other color palettes, direction reverses the palette 
  geom_sf_label(data = county_geo,aes(label = NAME), size = 2) +
  theme_minimal() +
  coord_sf(datum = st_crs(county_geo))
plot(map15_2)
```

[Q.5 Describe in 1-2 sentences - What do you notice about the spatial distribution of severe weather events based on this map?]{style="color: darkred;"} **\[5 points\]**

Notice how R automatically classified the sizes and colors. These are not always correct. We can also use algorithms to decide how to classify our size or colors. We can do this using the `classInt` library. For example, in the method used in the upcoming example, we use the 'natural breaks' (called 'jenks') classification method which divides the data based on natural classes. While we are not in a cartography class, below are a few good resources to under stand classification schemes, and mapping tips in general.

<https://www.axismaps.com/guide/data-classification>

<https://www.axismaps.com/guide>

<https://workshop.palaeoverse.org/NAPC2024/materials/4_visualization/practical-spatial.html>

```{r}
library(classInt) # load library

# Now we tell R how to 'cut up' or classify our data
# We provide the data column, how many bins/classes we want, and the style of classification
breaks_qt <- classIntervals(dat15$PROP_DAMAGE_USD, n = 5, style = "jenks", 
                            intervalClosure = c("right"))

# add the new cut values to dat15
dat15 <- dat15 |>
  mutate(damage_jenks5 = cut(PROP_DAMAGE_USD,breaks_qt$brks))

map15_3 <- ggplot() + 
  geom_sf(data = county_geo) +
  geom_sf(data = dat15,
          aes(size = damage_jenks5, color = damage_jenks5),
          alpha = 0.2) +
  scale_color_viridis_d(option = "rocket",
                        direction = -1 ) +
  geom_sf_label(data = county_geo,aes(label = NAME), size = 2) +
  theme_minimal() +
  coord_sf(datum = st_crs(county_geo)) +
  ggtitle("SC Extreme Events in 2015") # add title
plot(map15_3)
```

Notice how the map looks now with the classification schemes. The `help()` function in R is very useful to get an idea for how you need to write your code and what inputs you may need. So for example if you want to know want to know more about the function `geom_sf()`, you could type `help(geom_sf)` or `?geom_sf`

[Q.6 Now modifying the above code:]{style="color: darkred;"}

1\) [create your own map to look at the distribution of severe weather events in SC in 2024]{style="color: darkred;"}. **\[5 points\]**

2\) [in 1-2 sentences, describe how the spatial distribution of severe events is different 2024 compared 2015]{style="color: darkred;"} .**\[5 points\]**

```{r}
# Your code for Q6 1) here
```

One issue with plotting all events is that it might be misleading since different events have different geographies and magnitudes.

[Q.7 Create 2 maps, 1 for 2015, and 1 for 2024 showing the spatial distribution of extreme events categorized by the TYPE of EVENT, and by Property Damage.]{style="color: darkred;"} **\[10 points\]**

If you are curious about using different color schemes in R - here is a good resource.

<https://datavizf24.classes.andrewheiss.com/resource/colors.html>

```{r}
# your code here
#Hint: the code will be very similar to your previous map code. The main difference will be that you will need to assign a different variable (capturing event types) to the 'color =' argument.
#additionally change alpha to 0.5 and change the viridis option "magma"


```

## 3 Descriptive Spatial Statistics and Visualizations

While the maps are a good start, we can see that the story becomes complicated due to the large number of points on our map. Your task is to use descriptive spatial statistics and visualizations to answer the questions in this section.

To help you, I have already created all the descriptive spatial statistic functions for you. To load those functions in R, run the following code

```{r}
source("code/descriptive_geo_functions.R")
```

You will notice your Environment window (top right window) now has 6 functions that are loaded:

a\. `calc_mc()` - calculate mean center

b\. `calc_wmc()` - weighted mean center

c\. `calc_medc()` - median center

d\. `calc_weighted_medc()` - weighted median center

e\. `calc_weighted_sd()` - weighted standard distance

f\. `calc_sde_sf()` - weighted standard deviational ellipse

Below are 2 examples of calculating mean center and weighted mean center.

```{r}
mc <- calc_mc(dat15)
wmc <- calc_wmc(dat15,dat15$DEATHS_DIRECT)

```

Using appropriate statistics to answer the following questions. Use code and maps as needed to demonstrate your answer. [**Then in 1-2 sentences interpret your results**]{.underline}. Note that I do not care how you organize your code, maps, answers as long as you demonstrate evidence of how you reached the answers. I also encourage you to use `ggplotly()` for interactive plotting tool

[Q 8) How has the average geographic center for all severe weather events shifted in 2024, compared to 2025.]{style="color: darkred;"} **\[10 points\]**

[Q 9) How has the average geographic center, accounting for the magnitude of property damage for all severe weather events shifted in 2024, compared to 2025.]{style="color: darkred;"}**\[10 points\]**

[Q 10) Accounting for the magnitude of property damage, how are the average geographic centers different for the top 3 most frequently occuring event types in 2024.]{style="color: darkred;"}**\[10 points\]**

[Q 11) How did the spatial distribution and mean centers of flash floods change between 2015 and 2024?]{style="color: darkred;"} **\[10 points\]**

[Q 12) What is the difference between the mean center, distribution, and shape of thunderstorm winds between 2015 and 2024?]{style="color: darkred;"} **\[10 points\]**

[Q 13) What is the difference between the mean center, distribution, and shape of flash floods and regular floods in 2015? Do you notice any differences in 2024?]{style="color: darkred;"} **\[10 points\]**

[Q 14) Using data from 2024: Suppose you want to create a centralized disaster hub to respond to flash floods effectively. To do so you, you are looking to create a hub that minimizes the distance to all flash flooding events. Where would that hub be located?]{style="color: darkred;"} **\[10 points\]**

[Q 15) Using the same data from 2024, you also want to ensure that you are able to make more trips efficiently to locations that may cause higher property damage. Where would that hub be located?]{style="color: darkred;"} **\[10 points\]**

[Q 16) Would that hub been located differently in 2024, as compared 2015?]{style="color: darkred;"} **\[10 points\]**

[Bonus Q **\[Worth upto 20 points\]**: In real life, a single centralized hub to respond to different types of disasters (flood vs hail vs tornado) may not be very efficient as each disaster has its own response specifics.]{style="color: darkred;"}

[A better way would be to have a centralized hub for each disaster, and then one single master hub that helps assist each hub! How can you use the descriptive spatial statistics that you have learned to decide to where located these disaster specific hubs and a master hub!]{style="color: darkred;"}

## 4 Final Project - Question and Data Formulation

For your final project, you should have a main research question that is related to an outcome, AND also focuses on investigating the role of the underlying process(es) that may have generated the geographic distribution of the outcome that you are interested in.

For example, related to this exercise, my final project question could be:

*1) Does higher social vulnerability in an area (cause/process) increase the risk of extreme weather events will higher property damage (outcome) in South Carolina?*

Or another very simple but answerable question could be:

*2) Does the type of extreme weather event such as flash flood vs tornado (outcome) influence how much property damage (outcome) counties in South Carolina may suffer.*

If you dont already have a question or data sets in mind, I suggest you do the following.

Choose a study area that you would like to focus on for your project. Search for both spatial and non-spatial data (that can be made spatial) for your study area that you can download. These data could be potentially related to the topic you chose for your 'Asking Spatial Questions' paper exercise - but if you do not want to pursue your idea further or if you cannot find data, you may choose a different topic). I have added a data resource page for you but let your imagination flow.

You can also use ChatGPT to help but please be mindful of the accuracy of what it helps you with. I would rather that you think of a questions first and have some potential data ideas in mind and then ask ChatGPT to help refine. Please DO NOT USER IT TO GENERATE IDEAS FOR YOU. If I have reason to believe you did use it for that purpose, I will ask you to explain your project and question verbally in detail to me. Instead if you are stuck, COME AND TALK TO ME and I will help you with your idea.

[Q 17) Describe your potential research question. If you are not sure you can also describe upto 2 potential research questions.]{style="color: darkred;"} **\[10 points\]**

[Q.18) Find and download at least 3-4 datasets that you think you might use in your project. Use the following guidance:]{style="color: darkred;"}

-   [The data must be downloadable (or retrieval from API such as census API)]{style="color: darkred;"}

-   [At least one of these datasets should be 'spatial' in nature. It could already be in a spatial format or should contain information that can be converted to a spatial object. (For example: a spreadsheet with latitude and longitude values)]{style="color: darkred;"}

-   [At the least, your data should contain some information about your potential outcome variable and some variable representing the process that you are interested in.]{style="color: darkred;"}

-   [At the least, one data should contain information about your study area boundary. (for example, SC county boundaries for 2024)]{style="color: darkred;"}

-   [Create a table that has basic information about the layers you locate, including filename, data description, and the data source (e.g., link to webpage or downloadable data). Below is an example with my hypothetical question. You can replace the table with your own data]{style="color: darkred;"} **\[15 points\]**

| Filename                     | Data Type                                              | Data Source                                                                           |
|------------------------------|--------------------------------------------------------|---------------------------------------------------------------------------------------|
| Storm_database_locations.csv | storm property damage and spatial locations of storms  | <https://www.ncei.noaa.gov/stormevents/ftp.jsp>                                       |
| county_geo                   | SC county boundaries                                   | Through TIGRIS package                                                                |
| SouthCarolina_county.csv     | South carolina county level Social Vulnerability Index | <https://www.atsdr.cdc.gov/place-health/php/svi/svi-data-documentation-download.html> |

[Q 19) Finally, Write down 4 descriptive questions that you can ask about your research question and your data. For example, take a look at the questions in Q4). Atleast 2 questions should be spatial in nature.]{style="color: darkred;"} **\[20 points\]**

[Q 20) In 1-2 sentences each answer the following:]{style="color: darkred;"}

-   [What did you like about this assignment?]{style="color: darkred;"}

-   [What did you most struggle with while working on the assignment ?]{style="color: darkred;"}

-   [Do you have any questions/concerns about your final project? I will go through them and help you wherever I can]{style="color: darkred;"} **\[10 points\]**

[Q 21) Make sure that your file renders into an html successfully. (You dont need to type an answer here).]{style="color: darkred;"} **\[ 10 points\]**
