---
title: "Homework 1: Once Upon a Stormy Evening"
author: "Varun Goel"
description: |
  Due Date: **September 25, Friday, 11:59 pm** \newline \newline
  **(150 points + 20 bonus points)**
execute:
  warning: false
  eval: false
  echo: true
date: "`r Sys.Date()`"
format: 
  html:
    toc: true
    toc-expand: 3
editor: visual
---

# Introduction

The purpose of this HW is for you to utilize your descriptive statistics and visualization skills to a real world dataset and problem. Here, you will describe how the spatial distribution and frequency of different severe weather events in South Carolina has changed over time using the tools that you have learned in class. Part A of the assignment (75 points) involves data visualization and descriptive analysis while Part B will extend the analysis to descriptive spatial statistics.

# Instructions (Read first)

- Go to `File –> Save as`, and save this as `hw1_answers.qmd`. Then you can use that file for filling in all your answers.

- However, you will notice that since lots of code sections have incomplete code, In the header of the `.qmd` file, we have a setting that reads as `eval: false` . This prevents the code from running. Hence, whenever you complete a code section, you could override that command by adding `#| eval: true` to each code chunk. This is done for the first two code chunks.

- Please make sure that your code is entered in your code block but your written answers are **OUTSIDE your codeblock**. Only put comments (using #) in your code to describe your code. Your answers (other than code) need to be written outside your code block.

- [**Generative AI Use:**]{.underline}

  - You are welcome to use ChatGPT or other generative AI for your assignment. I encourage to use it mainly for coding help as needed. However, I also encourage you to first try to code yourself based on what we have learned so that you can get some practice. Even if you can "pseudocode" (being able to tell R instructions in a human-readable way), you have done well. If you do use generative AI [**for any part of the assignment,**]{.underline} create a word doc titled `ai_prompts.docx`. For [**each prompt**]{.underline} that you provide to generative AI software, report the following in the document:

    - The Question number from your homework

    - Your exact prompt as inputted to the AI

    - The [RAW (]{.underline}unchanged) answer generated by the software

  - Note that even if you use AI, you are responsible for cross-checking your content and ensure you understand your outputs, and importantly, are able to explain it verbally if called upon.

  - If you do use AI, but do not disclose [**ALL**]{.underline} your prompts , [**I will consider it as plagiarism and take strict action including giving you a zero on your assignment and report it as an academic integrity incident**]{.underline}. There is a specific manner in which softwares like ChatGPT write code. If I detect those patterns or suspect you used AI but did not report it, I may ask you to solve the problem in front of me to ensure that you did the problem and not generative AI. Hence, to be safe, please report all your prompts even if you made little or extensive use of generative AI

- [**Working Together with other classmates**]{.underline}

  - [You are welcome to work with a classmate or discuss answers and code, but your homework should be your own work]{.underline}. If you do work with a classmate, please report in the document whom you worked with so that I give you the benefit of doubt in case there are similar answers with an another person's work. I encourage you to work together and problem solve together!!

- [**Getting help**]{.underline}

  - DONT WAIT till the Last Minute. Ask me for help, proactively. I am here for you.

  - Your readings/In-class labs are a great resource, and if you havent gotten a chance to go through them, this is your chance to do so!

## 1 Load packages and data

Wheneve you start an analysis, it is good practice to first install and load all your packages, and also import data that you may be using. We have already installed this packages in R Posit, but we will load them, and also import severe weather events data. Luckily for us, the **National Center for Environmental Information at NOAA** maintains a comprehensive database of severe weather events. You can read more about it here: <https://www.ncei.noaa.gov/stormevents/>

Run the following code below.

```{r}
#| eval: true
#load packages
library(tidyverse)
#library(sf) # for spatial data
#library(plotly) # For interactive plotting
#library(tidycensus) # for reading US census data in R
#library(tigris)# For downloading and reading US geographic administrative boundaries in R
#options(tigris_use_cache = TRUE)

### DONT TOUCH THIS NEXT LINE OF CODE ###
#census_api_key(key = "a0b70fd4d6380a164f8b90cef1e114f9a7ed6706", install = TRUE, overwrite = TRUE)
# load SC severe weather events 2015 and 2024
dat15 <- read_csv("data/sc_nhaz_15.csv")
dat24 <- read_csv("data/sc_nhaz_24.csv")
```

This is a cleaned up and truncated version of the actual dataset which is very messy. However, even though this data is 'clean', there can be several hidden issues. It is important to know what each variable represents and how it is coded. To know what these variables are, refer to the [**codebook (Page 1-6) in the docs folder**]{.underline}.

## 2 Examining your Data

### 2.1 Cleaning Data

Before we can answer our descriptive questions, we need to first examine the data to make sure it is ready for analysis.

[Q.1 Examine the following variables more closely. Explain what kind of variable (nominal, ordinal, interval, ratio) they [should be,]{.underline} and how [they are actually stored as]{.underline} in the 2024 dataset.]{style="color: darkred;"}**\[5 points\]**

1.  [EVENT_TYPE]{style="color: darkred;"}

2.  [DEATHS_DIRECT]{style="color: darkred;"}

3.  [DAMAGE_PROPERTY]{style="color: darkred;"}

Now we will make some changes to our dataframes to get them analysis ready. For example, The original \`DAMAGE_PROPERTY\` variable stores values using suffixes such as \`K\` for thousands and \`M\` for millions. We need to convert these to actual digits for R to work with them for analysis. The following function has been created for you to convert those values to dollars. Run the code below to create the function named `convert_damage`

```{r}
#| eval: true
# Convert property damage values such as 35.00K, 1.80M, and 2.00B to dollars
convert_damage <- function(x) {
  x <- toupper(trimws(as.character(x)))
  
  result <- suppressWarnings(
    readr::parse_number(
      x,
      na = c("", "NA", "N/A")
    )
  )
  
  result <- dplyr::case_when(
    is.na(x) ~ NA_real_,
    x == "" ~ NA_real_,
    grepl("B$", x) ~ result * 1e9,
    grepl("M$", x) ~ result * 1e6,
    grepl("K$", x) ~ result * 1e3,
    TRUE ~ result
  )
  
  result
}

```

Now we will apply this function to our datasets along with some other changes.

[Q.2 Complete the missing code below to get your 2015 and 2024 datasets analysis ready:]{style="color: darkred;"} **\[5 points\]**

```{r}

# We will first create new variables or fix old variables
# create a new column that stores property damage values in USD
# also create a column that combines injuries, and deaths
# convert event types to factor class (nominal variable)

# Here is the code For the 2015 dataset:
dat15 <- dat15 |>
  mutate(
    PROP_DAMAGE_USD = convert_damage(_____________________), # appying the function
    ALL_INJ = INJURIES_DIRECT + INJURIES_INDIRECT,
    ALL_DEATH = DEATHS_DIRECT + DEATHS_INDIRECT,,
    EVENT_TYPE_F = as.factor(EVENT_TYPE)
  )

# Now do the same for 2024 data

dat24 <- "YOUR CODE"

# Now we will also combine the two years into one data frame to make it easy for graphing. This is also the essence of 'tidy data'. Since both the 2015 and 2024 data are about storms and we need them for analysis, it is better to have one single dataframe with another variable that lets us know which year the the event is from.

all_storms <- bind_rows(dat15, dat24) |>
  mutate(
    YEAR = factor(YEAR),
    EVENT_TYPE_F = fct_infreq(EVENT_TYPE_F) # arranges by frequency of events
  )

# In addition, add another variable called TOTAL_HARM that includes "all injuries and all deaths" to the all_storms dataframe.
all_storms <- all_storms |>
  ______________(
    TOTAL_HARM = _______________________
  )

```

Now, for this exercise we will only keep events which do not have missing property damage data.

```{r}
all_storms <- all_storms |>
  filter(!is.na(PROP_DAMAGE_USD))
```

[Q.3:How many observations and variables are there in our final all_storms dataframe? What does each observation in this dataframe represent?]{style="color: darkred;"} **\[3 points\]**

### 2.2 Descriptive Statistics

#### [Q.4:Counts and proportions for a categorical variable]{style="color: darkred;"}

- Lets get warmed up! First, lets create a [frequency table and a bar plot]{.underline} showing the number of records for each `EVENT_TYPE`. The code has been written here for you to run.

  ```{r}

  # Frequency table for event type
  event_counts <- all_storms |>
    count(EVENT_TYPE, name = "n") |> # count frequency of each event
    arrange(desc(n)) # sort by the column 'n' descending order (highest frequency to lowest)

  event_counts  # print the frequency table

  # Bar plot showing event counts
   plot_eve_count <- ggplot(data = event_counts, # the data frame being used
         aes(x = n, # x axis is the 'n' variable from frequency table
             y = fct_reorder(EVENT_TYPE, n) # y is each event type sorted by frequency (highest first)
             )) +
    geom_col(fill = "steelblue") + # specify color for graph
    labs( # specify meaningful labels
      title = "Frequency of Severe Weather Event Types",
      x = "Number of records",
      y = "Event type"
    ) +
    theme_minimal() # one of the aesthetics that is used to draw the plots
   
   plot_eve_count
  ```

- The above descriptives are not very useful because they merge the 2 years. Now, create a [frequency table and a bar plot]{.underline} showing the number of records for each `EVENT_TYPE` and by `YEAR`. A part of the code has been written for you, but you have to fill in the blanks. Hints are provided in comments **\[5 points\]**

  ```{r}

  event_counts <- ________ |> #dataframe to use
    count(__________, __________, name = "n") |> # frequency (n) by year and event type
    arrange(________, desc(n)) # sort by year first, and then by descending frequency (n)

  event_counts

  plot_eve_count <- ggplot( data = _______________,
    aes(
      x = ______________,
      y = forcats::fct_reorder(EVENT_TYPE, n),
      fill = factor(YEAR) # use different 'fills' (colors) for each year value
    )
  ) +
    geom_col(position = "dodge") + #need this to display bars side by side for each year
    labs(
      title = "__________________",
      x = "________________",
      y = "_______________",
      fill = "Year" # gives a 'legend' to the meaning of the colors
    ) +
    theme_minimal()

  plot_eve_count

  ```

- Now extend the table to include proportions of each event, and then create a proportions graph **\[5 points\]**

```{r}

event_props <- event_counts |>
  group_by(___________) |>
  mutate(
    proportion = __________________, # formula for proportion
    percent = 100 * proportion
  ) |>
  ungroup() |>
  arrange(___________, desc(_______)) # sort by year and proportion from highest to lowest

event_props

plot_eve_prop <- ## YOUR CODE for the plot HERE ###### 
 
plot_eve_prop

```

- In 2–3 sentences, answer the following: **\[5 points\]**

  - Which event types are most common overall?

  - Does the comparison between 2015 and 2024 change when proportions rather than raw counts are used?

  - Why might proportions be more appropriate than counts for comparing the two years?

#### [Q.5: Exploring the distribution of property damage]{style="color: darkred;"}

Now you will use histograms and a density plot sto examine the distribution of reported property damage. Here [we will focus only on events that reported atleast some property damage]{.underline}.

- First create a histogram and density plot of all reported property damage **\[2 points\]**

```{r}

# Histogram of property damage
hist_raw <- ggplot(all_storms |> filter(_____________________), 
       aes(x = PROP_DAMAGE_USD)) +
  geom_histogram(
    bins = 40,
    fill = "darkorange",
    color = "white"
  ) +
  scale_x_continuous(labels = scales::label_dollar()) + #
  labs(
    title = "Distribution of Reported Property Damage",
    x = "Property damage in dollars",
    y = "Number of records"
  ) +
  theme_minimal()

hist_raw

# Density plot on a logarithmic x-axis
density_raw <- ggplot(all_storms |> filter(________________________), 
       aes(x = PROP_DAMAGE_USD))+
  geom_density(fill = "purple", alpha = 0.5) + # alpha handles transparency
  scale_x_continuous(labels = scales::label_dollar()) +
  labs(
    title = "Distribution of Reported Property Damage",
    subtitle = "Logarithmic x-axis",
    x = "Property damage in dollars",
    y = "Density"
  ) +
  theme_minimal()

density_raw

```

- Describe the distribution using the following terms: **\[4 points\]**

  - Whether distribution appears symmetric or skewed? Why?

  - Whether the mean will be lesser, greater, or equal to the median? Why?

  - Whether you observe potential outliers in the distribution? Why or why not?

- Now you will run a second version of the histogram and density plots using a logarithmic x-axis.

```{r}

# Histogram of property damage using logarithmic axis
hist_log <- ggplot(all_storms |> filter(PROP_DAMAGE_USD > 0), 
       aes(x = PROP_DAMAGE_USD)) +
  geom_histogram(
    bins = 40,
    fill = "darkorange",
    color = "white"
  ) +
  scale_x_log10(labels = scales::label_dollar()) + #convert to logarithmic scale
  labs(
    title = "Distribution of Reported Property Damage",
    x = "Property damage in dollars",
    y = "Number of records"
  ) +
  theme_minimal()

hist_log

# Density plot on a logarithmic x-axis
density_log <- ggplot(
  all_storms |> filter(PROP_DAMAGE_USD > 0),
  aes(x = PROP_DAMAGE_USD)
) +
  geom_density(fill = "purple", alpha = 0.5) +
  scale_x_log10(labels = scales::label_dollar()) +
  labs(
    title = "Distribution of Reported Property Damage",
    subtitle = "Logarithmic x-axis",
    x = "Property damage in dollars",
    y = "Density"
  ) +
  theme_minimal()

density_log
```

- Briefly explain whether the Logarithmic scale in this case is a better option than the standard scale in observing distribution of property damage? Why or Why not? **\[3 points\]**
- Now consider the following scenarios, and explain which scale might be better to use in those scenarios and why? Note that neither scale is 'universally' correct and that the choice of one graph compared to the other depends on your purpose. When you answer the questions, think about these scenarios in general, and not based on what you see in the current dataset. **\[6 points\]**
  1.  A county emergency manager wants to know "**“What dollar amount of property damage is most typical for a reported severe-weather event?”**

  2.  A state emergency-management agency wants to identify the most financially damaging individual events and determine whether any observations are unusually extreme.

  3.  A public-information officer is preparing a report for residents who are unfamiliar with logarithmic scales. The report should answer: **“How much property damage was reported in each event?”** and the goal is clear communication rather than detailed statistical exploration. Which graph would be easier for the general public to interpret: the raw-scale or logarithmic-scale graph? Is the easier graph necessarily the most informative graph? Explain.

#### [Q.6: Comparing numerical distributions with boxplots]{style="color: darkred;"}

- Now, lets create a boxplots of property damage by event type for events that have experienced atleast some property damage. We will use a logarithmic y-axis so that smaller and larger damage values can be displayed together. **\[3 points\]**

```{r}

ggplot(
  all_storms |> filter(____________),
  aes(
    x = fct_reorder(EVENT_TYPE, PROP_DAMAGE_USD, .fun = median),# reorder the levels of the categorical variable EVENT_TYPE based on the median of PROP_DAMAGE_USD
    y = _____________________ 
  )
) +
  geom_boxplot(fill = "skyblue", outlier.color = "red") +
  scale_y_log10(labels = scales::label_dollar()) + #use log scale on y axis
  labs(
    title = "Property Damage by Event Type",
    x = "_________________",
    y = "___________________"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1) #stylistic adjustment for better fonts
  )

```

- Then create separate boxplots for 2015 and 2024 to compare property damage side by side for each event across years. **\[3 points\]**

```{r}
ggplot(
  all_storms |> filter(_________________________),
  aes(
    x = fct_reorder(EVENT_TYPE, PROP_DAMAGE_USD, .fun = median),
    y = ___________________,
    fill = _______________
  )
) +
  geom_boxplot(
    position = position_dodge(width = 0.8),# to ensure that boxplots can be shown across years
    outlier.color = "red"
  ) +
  scale_y_log10(labels = scales::label_dollar()) +
  labs(
    title = "Property Damage by Event Type and Year",
    x = "_______________",
    y = "________________",
    fill = "____________"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

- Based on the boxplot above, answer in 1 sentence each: **\[5 points\]**

  1.  Were there more outlier events in 2015 or 2024?
  2.  Which event types caused more damage in 2024 compared to 2015?
  3.  Which event types caused less damage in 2024 compared to 2015?
  4.  Which event type had the highest median value and in what year?
  5.  Which event types experienced an increase in spread in 2024 as compared to 2015?

#### [Q.6: Mean, median, and skewness]{style="color: darkred;"}

- Finally, calculate the mean, median, standard deviation, first quartile, and third quartile of property damage for each event type and year (that caused atleast some damage) in one single summary table **\[10 points\]**

  ```{r}
  damage_summary <- all_storms |> 
    filter(_____________________) |>
    group_by(____________, _________________________) |>
    summarise(
      n = n(),
      mean_damage = ___________________________,
      median_damage = __________________________,
      sd_damage = ________________________________,
      q1_damage = quantile(PROP_DAMAGE_USD, 0.25), # 1st quartile (25%)
      q3_damage = quantile(PROP_DAMAGE_USD, 0.75), # 3rd quartile (75%)
      iqr = ___________________________,
      max_damage = ________________________________________,
      .groups = "drop"
    ) |>
    mutate(
      mean_median_ratio = mean_damage / median_damage # calculate mean to meidan ratio
    ) |>
    arrange(YEAR, desc(mean_median_ratio))

  damage_summary
  ```

<!-- -->

- Based on the table, answer the following questions: **\[6 points\]**

  - For 2015 **`Flood`** events, is there a large difference in the mean property damage compared to the median? If so, what does this difference suggest about the distribution of property damage for these events? Which measure should be preferred when describing typical flooding damage in 2015? Why?

  - Take a look at the 2024 Hail events? Why might standard deviation be much larger as compared to IQR? Which might be a better statistic that may more realistically show the variation in property damage due to hail in 2024?

- The table and the boxplot summarize the same property-damage data. Compare the table with the boxplot and answer the following in **one short paragraph**: **\[5 points\]**

  1.  Identify one comparison that can be made reasonably well from the table alone.

  2.  Identify one important feature that is much easier to see in the boxplot than in the table.

  3.  Describe one example of a potential outlier visible in the boxplot.

  4.  Explain why the mean, median, standard deviation, IQR, and maximum value cannot completely replace a visualization.

  Your answer should refer to at least one specific event type and year.
