---
title: "Geographic Data Visualization"
author: "Varun Goel"
date: "`r Sys.Date()`"
format: html
editor: visual
---

# Introduction

The main purpose of this lab is to:

1\) Introduce to visualizing Geographic Data

2\) Implementing Descriptive Spatial Statistics in R

3\) Creating seamless text + code documents

## 1. Load Packages

```{r}
#| label: setup
#| include: false

# Set global knit options (optional)
knitr::opts_chunk$set(echo = T,warning = F,message = F)

# load packages
library(tidyverse)
library(sf)
library(DescTools) # for weighted median
library(spatstat) # for ellipse calculations
library(plotly) #  for interactive


```

## 2. Import Data

We will now load the data on deaths and pumps from Jon Snow's Cholera example

```{r}
#| label: data

# load deaths shapefile
deaths_shp <- st_read("data/deaths_nd_by_house.shp")

# read deaths data from csv
deaths_loc <- read_csv("data/deaths_nd_by_house.csv") |>
  st_as_sf(coords = c("COORD_X","COORD_Y"),
           crs = st_crs(deaths_shp)) |>
  filter(deaths > 0) # only look at houses with deaths

pumps_loc <- st_read("data/pumps.shp")
```

## 3. Describe (Map) the data

Now we will plot or map the data as we would in a GIS software. While Rstudio has the base `plot()` function that can be used to make a map, it is very limited and not very intuitive. Instead we will rely on the fantastic `ggplot` library (already loaded as part of the `tidyverse` package) on most of our data visualization.

I highly recommend you read Chapter 1 of the fantastic Data Science in R book to become more familiar with ggplot and data visualization in general. We will also look at concepts from Chapter 9 - 11 in due time.

<https://r4ds.hadley.nz/data-visualize.html>

### A simple basic dot/point map

Here is a simple basic point map using `ggplot()`.

```{r}
#| label: map1

map1 <- ggplot() + # tell R that you are creating a graphic
  geom_sf(data = deaths_loc) # tell R what GIS layer you are using.
plot(map1)
```

The great part here is that R knows that since this is a point object, it should plot this data as points. However, this map is basic.

### Proportional Symbol Map

Now we show the deaths, but we show it in a way where we can visually see locations with more vs less deaths. We will provide some aesthetics to our graphics using the `aes()` command.

```{r}
#| label: map2

map2 <- ggplot() +
  geom_sf(data = deaths_loc,
          aes(size = deaths), # change size of each point based on number of deaths
          color = "red", # assign red color to each point
          alpha = 0.2 # make each point 80% transparent to be able to see overlapping points
          ) +
  coord_sf(datum = st_crs(deaths_loc))
plot(map2) + 
  theme_minimal() # provide a more minimalistic theme
```

### Proportional Symbol Map + other features

Now lets also add the location of the pumps. Hopefully you will see the magic and ease of `ggplot()` here.

```{r}
#| label: map3

map3 <- ggplot() +
  geom_sf(data = deaths_loc,
          aes(size = deaths), # change size of each point based on number of deaths
          color = "red", # assign red color to each point
          alpha = 0.2 # make each point 80% transparent to be able to see overlapping points
          ) +
  geom_sf(data = pumps_loc,
          color="black", size=5, shape=8) +
  coord_sf(datum = st_crs(deaths_loc))

plot(map3) + 
  theme_minimal() # provide a more minimalistic theme
```

Can you see why we did not put the `size` argument in `aes()`?

you can find more information about different styles at <https://ggplot2.tidyverse.org/articles/ggplot2-specs.html> Another way to make this map is:

```{r}
#| label: map3_alt
map3 <- map2 +
  geom_sf(data = pumps_loc,
          color="black", size=3, shape=8)

plot(map3) + 
  theme_minimal() # provide a more minimalistic theme
```

## 4. Descriptive Geographic Statistics

Now lets look at how we can implement the various descriptive statistics we learned in class. Since there is no automatically written code here, we create our own function called `calc_mc`.

### Mean Geographic Center

Since there is no automatically written code here, we create our own function called `calc_mc`.

```{r}
# ------------------------
# 2. Mean geographic center (centroid of coords)
# ------------------------

calc_mc <- function(sf_dat){
  mean_center <- data.frame(
  X = mean(st_coordinates(sf_dat)[,1]),
  Y = mean(st_coordinates(sf_dat)[,2])
) %>% st_as_sf(coords = c("X", "Y"), crs = st_crs(sf_dat))
  return(mean_center)
}

mc_deaths <- calc_mc(deaths_loc)
mc_deaths
```

### Weighted Mean Center

We will plot these centers etc. later. Lets now look at the weighted mean center. Think about why we would want to calculate a weighted mean center here.

```{r}
# ------------------------
# 3. Weighted mean geographic center
# ------------------------

calc_wmc <- function(sf_dat,weight_col){
weighted_center <- data.frame(
  X = weighted.mean(st_coordinates(sf_dat)[,1], weight_col),
  Y = weighted.mean(st_coordinates(sf_dat)[,2], weight_col)
) %>% st_as_sf(coords = c("X", "Y"), crs = st_crs(sf_dat))
  return(weighted_center)
}

wmc_deaths <- calc_wmc(deaths_loc, deaths_loc$deaths)
wmc_deaths

```

### Standard Distance (weighted)

```{r}

calc_weighted_sd <- function(sf_dat, weight_col)
{
# formula: sqrt( sum(wi*((xi-x̄)^2 + (yi-ȳ)^2)) / sum(wi) )
  mean_X = weighted.mean(st_coordinates(sf_dat)[,1], weight_col)
  mean_Y = weighted.mean(st_coordinates(sf_dat)[,2], weight_col)

sdist <- sqrt(
  sum(weight_col * ((st_coordinates(sf_dat)[,1] - mean_X)^2 + (st_coordinates(sf_dat)[,2] - mean_Y)^2)) / sum(weight_col)
)

# Circle around weighted mean center
circle_coords <- function(center, r, n=100){
  angles <- seq(0, 2*pi, length.out=n)
  x <- center[1] + r*cos(angles)
  y <- center[2] + r*sin(angles)
  data.frame(X=x, Y=y)
}

circle_df <- circle_coords(c(mean_X, mean_Y), sdist) %>%
  st_as_sf(coords = c("X","Y"), crs=st_crs(sf_dat)) %>%
  summarise(geometry = st_combine(geometry)) %>%
  st_cast("POLYGON")

return(circle_df)

}

weighted_sd_sf <- calc_weighted_sd(deaths_loc,deaths_loc$deaths)
```

### Standard Deviational Ellipse

```{r}
calc_sde_sf <- function(sf_dat, weight_col)
{
  mean_X = weighted.mean(st_coordinates(sf_dat)[,1], weight_col)
  mean_Y = weighted.mean(st_coordinates(sf_dat)[,2], weight_col)
# Convert coords to matrix
X <- cbind(st_coordinates(sf_dat)[,1], st_coordinates(sf_dat)[,2])

# Weighted covariance
cov_w <- cov.wt(X, wt=weight_col)$cov

# Eigen decomposition (axes of ellipse)
eig <- eigen(cov_w)
radii <- sqrt(eig$values) * 2  # scale factor for 1 SD ellipse

theta <- seq(0, 2*pi, length=200)
ellipse <- t(t(eig$vectors %*% diag(radii)) %*% rbind(cos(theta), sin(theta)))
ellipse <- sweep(ellipse, 2, c(mean_X, mean_Y), FUN="+")

ellipse_df <- data.frame(X=ellipse[,1], Y=ellipse[,2]) %>%
  st_as_sf(coords=c("X","Y"), crs=st_crs(sf_dat)) %>%
  summarise(geometry = st_combine(geometry)) %>%
  st_cast("POLYGON")

return(ellipse_df)
}

sde_sf <- calc_sde_sf(deaths_loc,deaths_loc$deaths)
```

### 5. Plotting them all together

Now we will plot all the calculated statistics on the map

```{r}
map_stats <- map3 +
  geom_sf(data = mc_deaths, color="blue", size=3, shape=17) +
  geom_sf(data = wmc_deaths, color="darkgreen", size=3, shape=15) +
  geom_sf(data = weighted_sd_sf, fill=NA, color="black", linetype="dashed") +
  geom_sf(data = sde_sf, fill=NA, color="brown", linetype="solid") +
  labs(title="Geographic Descriptive Statistics",
       subtitle="Centers, Standard Distance, and Standard Deviational Ellipse",
       x="Easting", y="Northing") +
  theme_minimal() +
  coord_sf(datum = st_crs(deaths_loc))

plot(map_stats)
```

You can also make this plot interactive

```{r}
ggplotly(map_stats)
```

## Your Turn (Exercise)

You will now use the concepts learned in class to explore and interpret descriptive statistics on a slightly modified Jon Snow dataset. Before you do that I have also written the code for calculating the median and the weighted median center

```{r}
# ------------------------
# Median geographic center (component-wise median)
# ------------------------
calc_medc <- function(sf_dat){
median_center <- data.frame(
  X = median(st_coordinates(sf_dat)[,1]),
  Y = median(st_coordinates(sf_dat)[,2])
) %>% st_as_sf(coords = c("X", "Y"), crs = st_crs(sf_dat))
return(median_center)
}

# ------------------------
# Weighted median geographic center
# ------------------------
calc_weighted_medc <- function(sf_dat, weight_col)
{
weighted_median_center <- data.frame(
  X = Median(st_coordinates(sf_dat)[,1], w = weight_col),
  Y = Median(st_coordinates(sf_dat)[,2], w = weight_col)
) %>% st_as_sf(coords = c("X", "Y"), crs = st_crs(sf_dat))
return(weighted_median_center)
}

```

Now that you have all the functions necessary to run your analysis, your task is to complete the following:

1.  Read in the `deaths_modified.csv` file, assign the same crs as `deaths_loc` and only include households with deaths.

```{r}
# Your code for Q1 here
```

2.  Run the following descriptive statistics: mean center, weighted mean center, median center, weighted median center, standard distance, and standard deviational ellipse.

```{r}
# Your code for Q2 here
```

3.  Plot your map (both static and interactive) with the deaths and your statistics (no need to map out pumps)

```{r}
# your code for Q3 here
```

4.  Discuss and interpret your statistics. Which ones do you think are appropriate/inappropriate and why? Discuss in 3-4 sentences. (Note you dont need to add a code chunk here. Just type your answer in regular text)

5.  Render your file and make sure it successfully renders as an html.
