---
title: "Geog 531: Quantitative Methods in Geographic Research"
date: last-modified
date-format: medium
output:
  html_document:
    link-external-newwindow: true
    theme: flatly
    toc: yes
    toc_float: yes
    toc_depth: 3
---

------------------------------------------------------------------------

### Instructor Information

**Instructor:** Dr. Varun Goel

**Office:** Callcott 111

\
**Appointment link:** [Click Here](https://outlook.office.com/bookwithme/user/675339bb37584ebaa37d12437cb4f286@sc.edu/meetingtype/aKkUKwRTska6ndT42c8OWg2?anonymous&ep=mlink)

**Email:** <a href="mailto:varungoel@sc.edu">varungoel\@sc.edu</a>

------------------------------------------------------------------------

### Course Essentials

**Meeting times:** Monday, Wednesday *3:55 pm - 5:10pm*\
**Meeting location:** Callcott 302

**Student Support Hours:** Mon-Wed 2 pm - 3 pm

**Appointment link:** [Click Here](https://outlook.office.com/bookwithme/user/675339bb37584ebaa37d12437cb4f286@sc.edu/meetingtype/aKkUKwRTska6ndT42c8OWg2?anonymous&ep=mlink)

**\]GEOG 531 Discussion Google Sheet:** [Click Here soon](https://piazza.com/sc/spring2025/geog510/home)

------------------------------------------------------------------------

### Updated Syllabus

The latest class syllabus can downloaded [here](Geog531_FA25_QuantitativeMethods.pdf)

------------------------------------------------------------------------

### Updated Schedule

------------------------------------------------------------------------

Please note that this schedule is subject to change (specifically, lecture topics and due dates). Please check this page periodically in case there are any changes

```{r include = FALSE}
library(knitr)
library(kableExtra)
library(tidyverse)
library(magrittr)

## Read data
schedule <- read_csv("schedule.csv")
## Format date column
schedule$TUES %<>% as.Date("%m/%d/%y") %>% format("%m/%d/%y")
schedule$THUR %<>% as.Date("%m/%d/%y") %>% format("%m/%d/%y")
```

```{r echo = FALSE}
## No NAs
opts <- options(knitr.kable.NA = "")
## Print table
kable(schedule) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = T) %>%
  column_spec(1, width = "4em") %>%
  column_spec(2, width = "5.5em") %>%
  column_spec(3, width = "5.5em") %>%
  column_spec(5, width = "9em") %>%
  column_spec(6, width = "12.5em")
```
