
---
title: "Lab: Point & Area Pattern Analysis"
subtitle: "GEOG 531"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
  cache: true
editor: visual
---

## Learning objectives

By the end of this lab you will be able to:

1. Diagnose first‑ vs second‑order structure in a **point pattern** using kernel density, `G/F/J`, `K/L`, `g(r)`, and simulation envelopes.
2. Build **spatial weights** for areal data and compute **global** and **local** autocorrelation (Moran's I, LISA, Gi\*, Join Count).
3. Apply **Empirical Bayes** smoothing and explain **MAUP** effects.
4. Interpret maps and statistics in plain language.

## Setup

```{r}
#| label: setup
pkgs <- c(
  "sf","spdep","spData","tmap","dplyr",
  "spatstat.data","spatstat.geom","spatstat.explore"
)
need <- pkgs[!pkgs %in% installed.packages()[,"Package"]]
if (length(need) > 0) install.packages(need, repos = "https://cloud.r-project.org")
invisible(lapply(pkgs, library, character.only = TRUE))
set.seed(531)
```

::: callout-note
Submit an **HTML** or **PDF** rendering of this Quarto document with your narrative answers under each **Your Turn** prompt.
:::

## Part A · Point patterns (Lansing woods)

### A1. Explore the pattern & intensity

```{r}
data(lansing, package = "spatstat.data")
plot(lansing, main = "Lansing woods: tree locations (by species)")
```

**Your Turn A1 (2–3 sentences):** Describe any visible first‑order trend (non‑uniform density). Which areas are denser?

```{r}
# Kernel density with two bandwidths
bw1 <- spatstat.explore::bw.diggle(lansing)
kd1 <- spatstat.geom::density.ppp(lansing, sigma = bw1)
kd2 <- spatstat.geom::density.ppp(lansing, sigma = bw1 * 2)  # over-smooth for comparison
par(mfrow = c(1,2)); plot(kd1); plot(kd2); par(mfrow = c(1,1))
```

**Your Turn A2:** Which bandwidth better captures structure here, and why?

### A2. CSR diagnostics

```{r}
# Quadrat test
spatstat.explore::quadrat.test(lansing, nx = 4, ny = 4)
# Nearest-neighbor summaries
G <- spatstat.explore::Gest(lansing)
F <- spatstat.explore::Fest(lansing)
J <- spatstat.explore::Jest(lansing)
par(mfrow = c(1,3)); plot(G); plot(F); plot(J); par(mfrow = c(1,1))
```

**Your Turn A3:** Do G/F/J indicate clustering or inhibition relative to CSR? Explain briefly.

```{r}
# K/L with global simulation envelope under CSR
envK <- spatstat.explore::envelope(
  lansing, spatstat.explore::Kest,
  nsim = 199, global = TRUE, correction = "Ripley",
  simulate = expression(spatstat.random::rpoispp(intensity(lansing)))
)
plot(envK)
```

**Your Turn A4:** Over what distance ranges does the observed K fall outside the envelope? What does that imply?

### A3. Inhomogeneity & marks

```{r}
kd <- spatstat.geom::density.ppp(lansing, sigma = bw1)
K_inhom <- spatstat.explore::Kinhom(lansing, lambda = kd)
plot(K_inhom, main = "Inhomogeneous K(r)")
```

**Your Turn A5:** Compare K and K\_inhom. What changes in your interpretation?

```{r}
# Cross-type interaction (choose two species)
levels(spatstat.geom::marks(lansing))
K_boak_hick <- spatstat.explore::Kcross(lansing, i = "blackoak", j = "hickory")
plot(K_boak_hick, main = "Cross-K: blackoak vs hickory")
```

**Your Turn A6:** Is there evidence of attraction or repulsion between your chosen species?

---

## Part B · Area patterns (North Carolina SIDS)

### B1. Build weights & map the variable

```{r}
data(nc, package = "spData")
nc$SIDR79 <- (nc$SID79 / nc$BIR79) * 1000
nb_q <- spdep::poly2nb(nc, queen = TRUE)
lw_q <- spdep::nb2listw(nb_q, style = "W")
tmap::tmap_mode("plot")
tmap::tm_shape(nc) + tmap::tm_polygons("SIDR79", title = "SIDS rate (1979)")
```

**Your Turn B1:** Comment on variance instability (where denominators are small).

### B2. Global Moran’s I and sensitivity to weights

```{r}
spdep::moran.test(nc$SIDR79, lw_q)

# Try 'B' (binary) and kNN (k = 4)
lw_b <- spdep::nb2listw(nb_q, style = "B")
knn4 <- spdep::knn2nb(spdep::knearneigh(sf::st_coordinates(sf::st_centroid(nc)), k = 4))
lw_knn4 <- spdep::nb2listw(knn4, style = "W")
list(
  W = spdep::moran.test(nc$SIDR79, lw_q)$estimate["Moran I statistic"],
  B = spdep::moran.test(nc$SIDR79, lw_b)$estimate["Moran I statistic"],
  kNN4 = spdep::moran.test(nc$SIDR79, lw_knn4)$estimate["Moran I statistic"]
)
```

**Your Turn B2:** How do the choices of weights change the statistic and why might that be?

### B3. LISA clusters & multiple testing

```{r}
lm <- spdep::localmoran(nc$SIDR79, lw_q)
nc$Z_Ii <- lm[, "Z.Ii"]
nc$p_Ii <- 2*pnorm(-abs(nc$Z_Ii))
nc$p_adj <- p.adjust(nc$p_Ii, method = "BH")

lag_rate <- spdep::lag.listw(lw_q, nc$SIDR79)
m <- mean(nc$SIDR79); ml <- mean(lag_rate)
quad <- ifelse(nc$SIDR79 >= m & lag_rate >= ml, "High-High",
        ifelse(nc$SIDR79 <  m & lag_rate <  ml, "Low-Low",
        ifelse(nc$SIDR79 >= m & lag_rate <  ml, "High-Low", "Low-High")))
nc$cluster <- ifelse(nc$p_adj < 0.05, quad, "Not sig (BH)")

tmap::tm_shape(nc) + tmap::tm_polygons("cluster", title = "LISA clusters (FDR q<.05)")
```

**Your Turn B3:** Report the count of each cluster type and interpret one cluster in context.

### B4. Hotspot (Gi\*) & EB smoothing

```{r}
gi <- spdep::localG(nc$SIDR79, lw_q)
nc$GiZ <- as.numeric(gi)
tmap::tm_shape(nc) + tmap::tm_polygons("GiZ", title = "Gi* z-score")

eb <- spdep::EBlocal(nc$SID79, nc$BIR79)
nc$SIDR79_EB <- eb$est * 1000
tmap::tm_shape(nc) + tmap::tm_polygons("SIDR79_EB", title = "EB smoothed rate")
```

**Your Turn B4:** How do EB‑smoothed maps change your identification of hotspots?

### B5. MAUP demonstration

```{r}
coords <- sf::st_coordinates(sf::st_centroid(nc))
km <- kmeans(coords, centers = 20, nstart = 20)
nc$zone <- factor(km$cluster)

nc_agg <- nc |>
  dplyr::group_by(zone) |>
  dplyr::summarise(SID79 = sum(SID79, na.rm = TRUE), BIR79 = sum(BIR79, na.rm = TRUE)) |>
  dplyr::mutate(rate = (SID79/BIR79) * 1000)

nb_a <- spdep::poly2nb(nc_agg)
lw_a <- spdep::nb2listw(nb_a, style = "W")

c(
  Moran_I_counties = spdep::moran.test(nc$SIDR79, lw_q)$estimate["Moran I statistic"],
  Moran_I_zones    = spdep::moran.test(nc_agg$rate, lw_a)$estimate["Moran I statistic"]
)
```

**Your Turn B5:** What changed after aggregation? Why is it important to report the areal units and weights definition?

---

## Deliverables (what to submit)

1. Your rendered **HTML/PDF** lab with code, outputs, and short interpretations under each “Your Turn” item.
2. A brief (≤150 words) **reflection** on what surprised you or what you found most useful.

### Rubric (10 points total)

- Reproducibility & code quality (2)
- Point pattern analysis (3)
- Areal analysis (3)
- Interpretation & reflection (2)

## (Optional) Extensions

- Compare **Geary’s C** to Moran’s I.
- Try alternative bandwidth selectors (`bw.ppl`, `bw.scott`).
- Use **distance‑band** weights and re‑run LISA/Gi\*.
- Explore **cross‑L** for another pair of species in *lansing*.
```

