
---
title: "Point & Area Pattern Analysis"
subtitle: "GEOG 531 — Part I (Point patterns) & Part II (Area patterns)"
author: "Instructor: <your name>"
format:
  revealjs:
    slide-number: true
    progress: true
    hash: true
    toc: true
    toc-depth: 2
    transition: fade
    incremental: true
    code-overflow: wrap
    self-contained: true
    center: false
execute:
  echo: true
  warning: false
  message: false
  cache: true
date: "Updated: 2025-10-19"
---

```{r}
#| label: setup
#| include: false
# Core packages used across both parts
pkgs <- c(
  "sf","spdep","spData","tmap",
  "spatstat.data","spatstat.geom","spatstat.explore"
)
need <- pkgs[!pkgs %in% installed.packages()[,"Package"]]
if (length(need) > 0) install.packages(need, repos = "https://cloud.r-project.org")
invisible(lapply(pkgs, library, character.only = TRUE))
set.seed(531) # reproducibility for examples
```

## How to use this deck

- Designed for **two 75‑minute class meetings**.
- **Part I (Point patterns)**: first‑ & second‑order properties; CSR; summary functions (*G, F, J, K, L, g*); kernel density; inhomogeneity; bivariate interaction; simulation & envelopes.
- **Part II (Area patterns)**: choropleths & variance instability; spatial weights; global & local autocorrelation (Moran's I, Geary's C, Join Count); **LISA & Gi\***; Empirical Bayes smoothing; **MAUP**; brief on spatial regression.
- All examples run on **built‑in datasets** (no external downloads).

::: callout-note
**Tip for teaching**: Use the vertical navigation in Reveal.js to keep Part I and Part II separate in class.
:::

---

# Part I · Point pattern analysis (≈ 75 min)

## Recap & framing

- A **point pattern** is a set of event locations in a defined **window**.
- **First‑order**: variation in **intensity** (trend) across space.
- **Second‑order**: **interaction** among events (clustering, inhibition) beyond intensity.
- **CSR (Complete Spatial Randomness)** = homogeneous Poisson process.
  - *First‑order*: constant intensity.
  - *Second‑order*: no interaction.
- Key practicalities: **window definition**, units/CRS, duplicates, edge effects.

::: callout-tip
If the process is **inhomogeneous**, test interaction with inhomogeneous functions rather than K under CSR.
:::

## Getting a feel for the data

We'll use the classic **Lansing woods** tree dataset (*spatstat*).

```{r}
# Load a marked point pattern of tree species
data(lansing, package = "spatstat.data")
lansing
plot(lansing, main = "Lansing woods: tree locations (by species)")
```

## First‑order: intensity & kernel density

```{r}
# Bandwidth selection (one option among several)
bw_dig <- spatstat.explore::bw.diggle(lansing)
bw_dig
# Kernel density estimate image
kd <- spatstat.geom::density.ppp(lansing, sigma = bw_dig)
plot(kd, main = "Kernel density (bw = bw.diggle)"); points(lansing, pch = 20, cex = .3)
```

::: callout-note
Bandwidth matters: show a **too‑small** vs **too‑large** σ to illustrate under/over‑smoothing.
:::

## Quadrat counts test (first‑order)

```{r}
Q <- spatstat.geom::quadratcount(lansing, nx = 4, ny = 4)
plot(lansing, main = "Quadrat counts (4 × 4)"); plot(Q, add = TRUE, col = "red", cex = .9)
spatstat.explore::quadrat.test(lansing, nx = 4, ny = 4)
```

## Second‑order summaries: G, F, J

```{r}
G <- spatstat.explore::Gest(lansing)
F <- spatstat.explore::Fest(lansing)
J <- spatstat.explore::Jest(lansing)
par(mfrow = c(1,3))
plot(G, main = "Nearest-neighbor G(r)")
plot(F, main = "Empty space F(r)")
plot(J, main = "J(r) = (1-F)/(1-G)")
par(mfrow = c(1,1))
```

## Ripley's K & L; pair correlation g(r)

```{r}
K <- spatstat.explore::Kest(lansing, correction = c("border","Ripley"))
L <- spatstat.explore::Lest(lansing, correction = "Ripley")
g <- spatstat.explore::pcf(lansing)
par(mfrow = c(1,3))
plot(K, main = "K(r) with edge correction")
plot(L, main = "L(r) = sqrt(K/π)")
plot(g, main = "Pair correlation g(r)")
par(mfrow = c(1,1))
```

## Testing against CSR with simulation envelopes

```{r}
envK <- spatstat.explore::envelope(
  lansing, spatstat.explore::Kest,
  nsim = 99, global = TRUE, correction = "Ripley",
  simulate = expression(spatstat.random::rpoispp(intensity(lansing)))
)
plot(envK, main = "Global simulation envelope for K(r) under CSR")
```

::: callout-important
**Global vs pointwise envelopes**: emphasize family‑wise error and interpretation.
:::

## Inhomogeneity: trend & inhomogeneous K

```{r}
# Use the kernel intensity surface as proxy for λ(s)
K_inhom <- spatstat.explore::Kinhom(lansing, lambda = kd)
plot(K_inhom, main = "Inhomogeneous K(r)")
```

## Marked & multitype patterns (cross‑K)

```{r}
levels(spatstat.geom::marks(lansing))
K_boak_hick <- spatstat.explore::Kcross(lansing, i = "blackoak", j = "hickory")
plot(K_boak_hick, main = "Cross-K: blackoak vs hickory")
```

## Nearest‑neighbor based test (Clark–Evans)

```{r}
spatstat.explore::clarkevans.test(lansing)  # R < 1 → clustering; R > 1 → regularity
```

## Key pitfalls & remedies

- Edge effects → use **edge corrections** (border, isotropic, Ripley).
- Window misspecification → **sensitivity analysis**.
- Inhomogeneity → use **inhomogeneous summaries**.
- Multiple testing (many *r* values) → **global envelopes** or **Monte Carlo rank** tests.
- **Interpretation**: summary functions suggest patterns; they do not identify mechanisms.

---

# Part II · Area (lattice) pattern analysis (≈ 75 min)

## Areal data vs point processes

- Variables measured over **zones** (counties, tracts, grids).
- Beware **MAUP** (scale/zonation) and **ecological fallacy**.
- **Rates** are often unstable when denominators vary; use **Empirical Bayes** smoothing.

## Example dataset: North Carolina SIDS (spData::nc)

```{r}
data(nc, package = "spData")
nc$SIDR79 <- (nc$SID79 / nc$BIR79) * 1000  # infant deaths per 1,000 births
nc <- sf::st_transform(nc, 32119)          # projected CRS (units in meters/feet depending on EPSG)
tmap::tmap_mode("plot")
tmap::tm_shape(nc) + tmap::tm_polygons("SIDR79", title = "SIDS rate (1979)")
```

## Spatial weights: contiguity & kNN

```{r}
# Queen contiguity
nb_q <- spdep::poly2nb(nc, queen = TRUE)
lw_q <- spdep::nb2listw(nb_q, style = "W")
# Visualize neighbors
plot(sf::st_geometry(nc), border = "grey")
plot(nb_q, sf::st_coordinates(sf::st_centroid(nc)), add = TRUE)
```

::: callout-note
**Weight styles**: “W” (row‑standardized), “B” (binary), “S” (variance‑stabilizing). Try each and compare.
:::

## Global autocorrelation: Moran's I (permutation test)

```{r}
mi <- spdep::moran.test(nc$SIDR79, lw_q)
mi
mi_mc <- spdep::moran.mc(nc$SIDR79, lw_q, nsim = 999)
mi_mc
```

## Local Moran (LISA) & cluster map

```{r}
lm <- spdep::localmoran(nc$SIDR79, lw_q)
nc$Ii   <- lm[, "Ii"]
nc$Z_Ii <- lm[, "Z.Ii"]               # z-scores
nc$p_Ii <- 2*pnorm(-abs(nc$Z_Ii))     # two-sided p
lag_rate <- spdep::lag.listw(lw_q, nc$SIDR79)
m <- mean(nc$SIDR79, na.rm = TRUE); ml <- mean(lag_rate, na.rm = TRUE)
nc$quad <- ifelse(nc$SIDR79 >= m & lag_rate >= ml, "High-High",
           ifelse(nc$SIDR79 <  m & lag_rate <  ml, "Low-Low",
           ifelse(nc$SIDR79 >= m & lag_rate <  ml, "High-Low", "Low-High")))
nc$quad_sig <- ifelse(nc$p_Ii < 0.05, nc$quad, "Not sig")

tmap::tm_shape(nc) +
  tmap::tm_polygons("quad_sig", title = "LISA clusters (p < .05)")
```

::: callout-caution
Local tests involve **many comparisons**. Adjust *p*‑values (e.g., `p.adjust(..., method="BH")`) or use **FDR** control.
:::

## Hot/cold spots: Getis–Ord Gi\*

```{r}
gi <- spdep::localG(nc$SIDR79, lw_q)
nc$GiZ <- as.numeric(gi)
tmap::tm_shape(nc) + tmap::tm_polygons("GiZ", title = "Gi* z-score")
```

## Binary pattern: Join count

```{r}
nc$HighSID <- nc$SIDR79 > median(nc$SIDR79, na.rm = TRUE)
spdep::joincount.test(nc$HighSID, lw_q, nsim = 999)
```

## Empirical Bayes smoothing of unstable rates

```{r}
eb <- spdep::EBlocal(nc$SID79, nc$BIR79)  # returns matrix with $raw, $est, etc.
nc$SIDR79_EB <- eb$est * 1000  # EB 'est' is usually per 1; scale per 1,000 births
# Compare raw vs EB smoothed
tmap::tm_shape(nc) + tmap::tm_polygons("SIDR79", title = "Raw rate (per 1,000)") +
tmap::tm_shape(nc) + tmap::tm_borders(col = NA) +
tmap::tm_shape(nc) + tmap::tm_polygons("SIDR79_EB", title = "EB smoothed (per 1,000)")
```

## MAUP demo: change the zoning & recompute Moran's I

```{r}
# Cluster county centroids into 20 regions and aggregate
coords <- sf::st_coordinates(sf::st_centroid(nc))
km <- kmeans(coords, centers = 20, nstart = 20)
nc$zone <- factor(km$cluster)

nc_agg <- nc |>
  dplyr::group_by(zone) |>
  dplyr::summarise(
    SID79 = sum(SID79, na.rm = TRUE),
    BIR79 = sum(BIR79, na.rm = TRUE)
  ) |>
  dplyr::mutate(rate = (SID79 / BIR79) * 1000)

nb_a <- spdep::poly2nb(nc_agg, queen = TRUE)
lw_a <- spdep::nb2listw(nb_a, style = "W")
mi_orig <- spdep::moran.test(nc$SIDR79, lw_q)$estimate["Moran I statistic"]
mi_agg  <- spdep::moran.test(nc_agg$rate, lw_a)$estimate["Moran I statistic"]
cat("Moran's I (original counties):", round(mi_orig, 3), "\n")
cat("Moran's I (aggregated 20 zones):", round(mi_agg, 3), "\n")
```

::: callout-important
**Takeaway**: Spatial statistics can **change with scale and zoning**. Always report how weights and units were defined.
:::

## Where to next? (very brief)

- Spatial regression (SLX, SAR, SEM, CAR), spatial filtering, multiscale models.
- Space–time extensions: Knox/SaTScan for points; ST‑LISA for areal series.
- Reproducibility: `quarto`, `targets`/`drake`, pinned datasets.

## Summary

- **Point patterns**: use CSR and second‑order summaries wisely; check inhomogeneity.
- **Area patterns**: choose weights carefully; use LISA/Gi\* for local structure; smooth unstable rates; be explicit about MAUP.
- **Interpretation** matters more than the statistic: tie results back to a process story.
