---
title: "Exploratory Spatial Data Analysis"
subtitle: "Point Pattern and Area Pattern Analysis"
author: "Varun Goel"
institute: ""
date: 2026-09-22
date-format: medium
format:
  revealjs:
    theme: [default, ../../assets/style.scss]
    slide-number: true
    chalkboard: true
    logo: logo.png
    footer: "Quantitative Methods in Geographic Research"
    incremental: true
    smaller: true
    scrollable: false
---

# Lecture 1: Introduction to ESDA & Point Pattern Analysis {background-color="#003366"}

## Module Overview {.smaller}

### Course Structure (3 Lectures × 75 min)

**Lecture 1**: Foundations & Point Patterns\
- Introduction to Exploratory Spatial Data Analysis (ESDA) - Patterns vs. Processes - First-order vs. Second-order Effects - Nearest Neighbor Analysis

**Lecture 2**: Point Pattern Analysis (continued)\
- Quadrat Analysis - Comparing Methods - Applications & Case Studies

**Lecture 3**: Area Pattern Analysis\
- Join Count Analysis - Moran's I (Global & Local) - Applications to Areal Data

------------------------------------------------------------------------

## What is Exploratory Spatial Data Analysis?

::::: columns
::: {.column width="50%"}
### Core Concepts

- **Exploratory**: Initial investigation of spatial data
- **Visual**: Maps, plots, graphs
- **Statistical**: Quantitative measures
- **Hypothesis-generating**: Not confirmatory
:::

::: {.column width="50%"}
### Key Questions

1.  Is there spatial pattern?
2.  Where are the patterns?
3.  Are patterns clustered or dispersed?
4.  Are patterns random or systematic?
:::
:::::

::: callout-important
ESDA reveals structure before formal modeling
:::

------------------------------------------------------------------------

## Types of Spatial Data

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
library(ggplot2)
library(gridExtra)

# Point pattern
set.seed(123)
p1 <- ggplot(data.frame(x = runif(50, 0, 10), y = runif(50, 0, 10)), 
             aes(x, y)) +
  geom_point(size = 3, color = "blue") +
  theme_minimal() +
  ggtitle("Point Pattern Data") +
  labs(subtitle = "Disease cases, tree locations, crime incidents")

# Areal data
p2 <- ggplot(data.frame(x = c(0,1,1,0,0, 1,2,2,1,1, 0,1,1,0,0),
                        y = c(0,0,1,1,0, 0,0,1,1,0, 1,1,2,2,1),
                        group = rep(1:3, each=5),
                        value = c(20, 45, 30)), 
             aes(x, y, fill = factor(value), group = group)) +
  geom_polygon(color = "black", linewidth = 1) +
  scale_fill_brewer(palette = "Blues") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  ggtitle("Areal Data") +
  labs(subtitle = "Census tracts, counties, regions", fill = "Value")

grid.arrange(p1, p2, ncol = 2)
```

------------------------------------------------------------------------

## Patterns vs. Processes {.smaller}

::::: columns
::: {.column width="50%"}
### Pattern

- **What we observe**
- The spatial arrangement
- Descriptive
- Examples:
  - Clustered disease cases
  - Dispersed fire stations
  - Random plant distribution
:::

::: {.column width="50%"}
### Process

- **Why it exists**
- The generating mechanism
- Explanatory
- Examples:
  - Disease transmission
  - Planning optimization
  - Seed dispersal mechanisms
:::
:::::

::: callout-warning
## Critical Distinction

**Multiple processes can produce similar patterns!**\
**Same process can produce different patterns!**
:::

::: notes
Emphasize: We analyze patterns to infer processes, but this inference is not straightforward. Different underlying mechanisms can create similar-looking spatial arrangements.
:::

------------------------------------------------------------------------

## Pattern-Process Examples

### Urban Crime Distribution

**Observed Pattern**: Crime clustered in certain neighborhoods

**Possible Processes**:

1.  **Social disorganization** (community characteristics)
2.  **Routine activity theory** (opportunity structures)
3.  **Broken windows** (environmental cues)
4.  **Repeat victimization** (contagion effects)

::: callout-note
Same pattern, different theoretical explanations!\
Need theory + statistical analysis + domain knowledge
:::

------------------------------------------------------------------------

## First-Order vs. Second-Order Effects {.smaller}

### First-Order Effects (Intensity/Density)

- **Definition**: Variation in the *average* number of events across space
- **What it captures**: Spatial trend, large-scale variation
- **Example**: More disease cases in densely populated areas

### Second-Order Effects (Dependence/Interaction)

- **Definition**: Correlation between events at *different locations*
- **What it captures**: Local clustering, spatial dependence
- **Example**: Disease cases clustered due to contagion (beyond population density)

::: callout-tip
## Why This Matters

Different analysis methods target different effects!\
Understanding this helps you choose the right tool.
:::

------------------------------------------------------------------------

## Visual Comparison: First vs. Second Order

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 4
set.seed(456)

# First order - gradient
x_grad <- runif(200, 0, 10)
y_grad <- runif(200, 0, 10)
# More points in upper right
prob_grad <- x_grad * y_grad
keep_grad <- runif(200) < (prob_grad / max(prob_grad))
df_first <- data.frame(x = x_grad[keep_grad], y = y_grad[keep_grad])

# Second order - clusters
n_clusters <- 4
centers <- data.frame(x = c(2, 8, 3, 7), y = c(2, 8, 7, 3))
points_per_cluster <- 30
df_second <- data.frame()
for(i in 1:n_clusters) {
  cluster_points <- data.frame(
    x = rnorm(points_per_cluster, centers$x[i], 0.5),
    y = rnorm(points_per_cluster, centers$y[i], 0.5)
  )
  df_second <- rbind(df_second, cluster_points)
}

# Random for comparison
df_random <- data.frame(x = runif(120, 0, 10), y = runif(120, 0, 10))

p1 <- ggplot(df_random, aes(x, y)) +
  geom_point(size = 2, alpha = 0.6, color = "darkgreen") +
  theme_minimal() +
  ggtitle("Random (No Effects)") +
  xlim(0, 10) + ylim(0, 10)

p2 <- ggplot(df_first, aes(x, y)) +
  geom_point(size = 2, alpha = 0.6, color = "blue") +
  theme_minimal() +
  ggtitle("First-Order (Density Gradient)") +
  xlim(0, 10) + ylim(0, 10)

p3 <- ggplot(df_second, aes(x, y)) +
  geom_point(size = 2, alpha = 0.6, color = "red") +
  theme_minimal() +
  ggtitle("Second-Order (Clustering)") +
  xlim(0, 10) + ylim(0, 10)

grid.arrange(p1, p2, p3, ncol = 3)
```

------------------------------------------------------------------------

## First-Order Effects in Detail

### Characteristics

- Large-scale spatial trends
- "Density surface" varies across study area
- Related to environmental/contextual factors

### Examples

- Population density → disease incidence
- Elevation → vegetation types\
- Distance from CBD → housing prices

### Analysis Methods

- Kernel density estimation
- Trend surface analysis
- **Quadrat analysis** (density variation)

------------------------------------------------------------------------

## Second-Order Effects in Detail

### Characteristics

- **Spatial dependence**: "Everything is related, but nearby things more so" (Tobler's Law)
- Local clustering beyond overall density
- Point-to-point interactions

### Examples

- Contagious disease spread (beyond density)
- Tree competition effects
- Social influence on voting patterns

### Analysis Methods

- **Nearest Neighbor Analysis** (spacing patterns)
- K-functions, L-functions
- Spatial autocorrelation (for areal data)

------------------------------------------------------------------------

## Why the Distinction Matters {.smaller}

::: callout-important
## Interpretation Challenge

A clustered pattern could result from:

1.  **First-order effects only**: Varying density due to environmental factors
2.  **Second-order effects only**: Uniform density but local interactions
3.  **Both**: Density variation AND clustering within high-density areas
:::

**Implication for Analysis:** - Need appropriate null models - Consider study area boundaries - Think about scale of analysis - Use complementary methods

------------------------------------------------------------------------

# Point Pattern Analysis {background-color="#004d00"}

## Introduction to Point Patterns

### What Are Point Patterns?

**Definition**: Spatial arrangement of discrete events/objects represented as points

**Characteristics**: - **Location**: $(x, y)$ coordinates - **Marked**: May have attributes (size, type, time) - **Study area**: Defined region of interest

### Examples

- Disease cases (lat/lon of residences)
- Tree locations in a forest
- Crime incidents
- Fire/emergency service facilities

------------------------------------------------------------------------

## Goals of Point Pattern Analysis

### Primary Objectives

1.  **Describe** the spatial arrangement
    - Is it clustered, dispersed, or random?
2.  **Quantify** the degree of clustering/dispersion
    - How clustered/dispersed?
3.  **Test** against theoretical distributions
    - Is the pattern significantly different from random?
4.  **Infer** underlying processes
    - What mechanism generated this pattern?

------------------------------------------------------------------------

## Complete Spatial Randomness (CSR) {.smaller}

### The Null Hypothesis

**CSR Assumptions**:

1.  **Homogeneous Poisson process**
2.  **Constant intensity** $\lambda$ (events per unit area)
3.  **Independence**: Location of one event doesn't affect others
4.  **Random**: Each location equally likely

### Implications Under CSR

- Number of events in any region follows Poisson distribution
- Event locations uniformly distributed
- No clustering, no dispersion
- **This is our baseline for comparison**

::: callout-note
Most real-world patterns deviate from CSR!\
That's what makes spatial analysis interesting.
:::

------------------------------------------------------------------------

## Departures from CSR

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 3.5
set.seed(789)

# Clustered
df_clust <- data.frame()
for(i in 1:5) {
  center <- c(runif(1, 1, 9), runif(1, 1, 9))
  pts <- data.frame(
    x = rnorm(20, center[1], 0.6),
    y = rnorm(20, center[2], 0.6)
  )
  df_clust <- rbind(df_clust, pts)
}

# Random
df_rand <- data.frame(x = runif(100, 0, 10), y = runif(100, 0, 10))

# Dispersed (use regular grid with small jitter)
grid_pts <- expand.grid(x = seq(1, 9, length.out = 10), 
                        y = seq(1, 9, length.out = 10))
df_disp <- data.frame(
  x = grid_pts$x + rnorm(100, 0, 0.2),
  y = grid_pts$y + rnorm(100, 0, 0.2)
)

p1 <- ggplot(df_clust, aes(x, y)) +
  geom_point(size = 2.5, color = "red", alpha = 0.6) +
  theme_minimal() + ggtitle("CLUSTERED") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
  xlim(0, 10) + ylim(0, 10)

p2 <- ggplot(df_rand, aes(x, y)) +
  geom_point(size = 2.5, color = "blue", alpha = 0.6) +
  theme_minimal() + ggtitle("RANDOM (CSR)") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
  xlim(0, 10) + ylim(0, 10)

p3 <- ggplot(df_disp, aes(x, y)) +
  geom_point(size = 2.5, color = "darkgreen", alpha = 0.6) +
  theme_minimal() + ggtitle("DISPERSED") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
  xlim(0, 10) + ylim(0, 10)

grid.arrange(p1, p2, p3, ncol = 3)
```

**Clustered**: $R < 1$ (points closer than expected)\
**Random**: $R \approx 1$ (CSR)\
**Dispersed**: $R > 1$ (points farther apart than expected)

------------------------------------------------------------------------

## Nearest Neighbor Analysis (NNA) {.smaller}

### Concept

**Measures the spacing between points**

- For each point, find distance to its nearest neighbor
- Calculate mean nearest neighbor distance ($\overline{NND}$)
- Compare to expected distance under CSR

### Key Advantage

- Simple to calculate and interpret
- Sensitive to second-order clustering/dispersion

### Limitation

- Reduces all spatial information to a single number
- Sensitive to boundary effects
- Only uses nearest neighbor (ignores other distances)

------------------------------------------------------------------------

## NNA: The Mathematics

### Observed Mean Nearest Neighbor Distance

$$\overline{NND} = \frac{\sum NND}{n}$$

where $n$ = number of points

### Expected Distance Under CSR

$$\overline{NND}_E = \frac{1}{2\sqrt{Density}} = \frac{1}{2\sqrt{n/A}}$$

where:\
- $n$ = number of points\
- $A$ = area of study region\
- Density = $n/A$ (points per unit area)

------------------------------------------------------------------------

## NNA: R-Statistic {.smaller}

### Standardized Nearest Neighbor Index

$$R = \frac{\overline{NND}}{\overline{NND}_E}$$

### Interpretation

| R Value     | Pattern Type           | Interpretation                   |
|-------------|------------------------|----------------------------------|
| $R = 0$     | **Maximum clustering** | All points at same location      |
| $0 < R < 1$ | **Clustered**          | Points closer than random        |
| $R = 1$     | **Random (CSR)**       | Points follow CSR                |
| $R > 1$     | **Dispersed**          | Points farther apart than random |
| $R = 2.149$ | **Maximum dispersion** | Perfect hexagonal spacing        |

------------------------------------------------------------------------

## NNA: Statistical Significance

### Standard Error

$$\sigma_{NND} = \frac{0.26136}{\sqrt{n(Density)}}$$

### Test Statistic (Z-score)

$$Z_R = \frac{\overline{NND} - \overline{NND}_E}{\sigma_{NND}}$$

### Decision Rule

- If $|Z_R| > 1.96$ → Reject CSR at $\alpha = 0.05$
- Positive $Z$: significantly dispersed
- Negative $Z$: significantly clustered

------------------------------------------------------------------------

## NNA: Worked Example from Reading {.smaller}

**Toronto Community Services** (Police, Fire, Elementary Schools, Voting)

| Facility | n   | Density | $\overline{NND}$ | $\overline{NND}_E$ | R    | Z      | p-value |
|---------|---------|---------|---------|---------|---------|---------|---------|
| Police   | 20  | 0.0312  | 2.08             | 2.83               | 1.28 | 2.42   | 0.0153  |
| Fire     | 20  | 0.0312  | 2.08             | 2.83               | 1.28 | 2.42   | 0.0153  |
| Schools  | 20  | 0.7308  | 0.59             | \-                 | 0.84 | -12.23 | 0.0000  |
| Voting   | 20  | 24.4351 | 0.27             | \-                 | 0.84 | \-     | 0.0153  |

### Interpretations

- **Police & Fire**: Dispersed (planned spacing)
- **Schools**: Clustered (population density)
- **Voting**: Highly clustered

------------------------------------------------------------------------

## NNA in R: Basic Implementation

```{r}
#| echo: true
#| eval: false

# Using spatstat package
library(spatstat)

# Create point pattern object
pp <- ppp(x = coords$x, y = coords$y, 
          window = owin(c(0, 100), c(0, 100)))

# Calculate nearest neighbor distances
nnd <- nndist(pp)
mean_nnd <- mean(nnd)

# Expected under CSR
n <- npoints(pp)
area <- area(Window(pp))
density <- n / area
expected_nnd <- 1 / (2 * sqrt(density))

# R-statistic
R <- mean_nnd / expected_nnd

# Z-test
se_nnd <- 0.26136 / sqrt(n * density)
z_score <- (mean_nnd - expected_nnd) / se_nnd
p_value <- 2 * pnorm(-abs(z_score))
```

------------------------------------------------------------------------

## Boundary Effects in NNA

::: callout-warning
## The Problem

Points near study area boundary have truncated neighborhoods!
:::

### Types of Bias

- **Edge effects**: Nearest neighbors may lie outside study area
- **Guard area**: Use buffer zone (exclude boundary points)
- **Toroidal correction**: "Wrap around" boundaries
- **Border correction methods**: Weight by probability of inclusion

### Practical Recommendation

- Report study area definition
- Use correction methods when available
- Interpret carefully for irregular boundaries
- Consider sensitivity analysis

------------------------------------------------------------------------

## When to Use NNA

### Appropriate Situations ✓

- Relatively homogeneous study area
- Interest in overall spacing pattern
- Quick exploratory analysis
- Second-order effects (clustering)

### Less Appropriate ✗

- Strong first-order effects (use quadrat)
- Multiple scales of clustering
- Need to identify cluster locations
- Very irregular boundaries

::: callout-tip
**Best Practice**: Use NNA in combination with other methods (e.g., quadrat analysis, visual inspection)
:::

------------------------------------------------------------------------

# Summary: Lecture 1 {background-color="#660000"}

## Key Concepts Covered

1.  **ESDA Framework**
    - Exploratory approach to spatial data
2.  **Patterns vs. Processes**
    - Observable vs. generative mechanisms
3.  **First vs. Second-Order Effects**
    - Density variation vs. spatial dependence
4.  **Point Pattern Analysis Introduction**
    - Complete Spatial Randomness (CSR)
5.  **Nearest Neighbor Analysis**
    - Measurement, interpretation, significance testing

------------------------------------------------------------------------

## Before Next Class

### Required Reading

- Review Lembo Chapter 14 (sections 14.1-14.2)
- Focus on examples and interpretation

### Think About

- What spatial patterns exist in your research area?
- Are they first-order or second-order effects?
- What processes might generate these patterns?

### Next Lecture

- Quadrat Analysis
- Variance-Mean Ratio (VMR)
- Comparing NNA and Quadrat approaches
- More case studies

------------------------------------------------------------------------

# Lecture 2: Point Pattern Analysis (Continued) {background-color="#003366"}

## Lecture 1 Recap {.smaller}

::::: columns
::: {.column width="50%"}
### Conceptual Foundations

- **ESDA**: Exploratory spatial data analysis
- **Pattern vs. Process**: Observable vs. mechanism
- **First-order**: Density/intensity variation
- **Second-order**: Spatial dependence
- **CSR**: Complete Spatial Randomness baseline
:::

::: {.column width="50%"}
### Nearest Neighbor Analysis

- Measures point **spacing**
- Mean nearest neighbor distance
- R-statistic: $\overline{NND} / \overline{NND}_E$
- Sensitive to **second-order effects**
- Good for overall pattern assessment
:::
:::::

### Today's Focus

**Quadrat Analysis** - A complementary approach focusing on **density variation**

------------------------------------------------------------------------

## Quadrat Analysis: Introduction {.smaller}

### Core Concept

**Divide study area into cells (quadrats) and count points in each cell**

### What It Measures

- **Variability in cell frequencies** (not spacing)
- Primarily **first-order effects** (density variation)
- Pattern of points across the study area

### Key Difference from NNA

- NNA: Point-to-point distances (spacing)
- Quadrat: Point counts per area unit (density)

::: callout-note
Quadrat analysis is analogous to a **frequency distribution** for spatial data
:::

------------------------------------------------------------------------

## Quadrat Analysis: Visual Concept

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 5
set.seed(101)

# Create three patterns
# Clustered
centers_c <- data.frame(x = c(2, 8, 2, 8), y = c(2, 2, 8, 8))
df_c <- data.frame()
for(i in 1:4) {
  pts <- data.frame(
    x = rnorm(15, centers_c$x[i], 0.8),
    y = rnorm(15, centers_c$y[i], 0.8)
  )
  df_c <- rbind(df_c, pts)
}

# Random
df_r <- data.frame(x = runif(60, 0, 10), y = runif(60, 0, 10))

# Dispersed
grid_base <- expand.grid(x = seq(0.5, 9.5, length.out = 8),
                         y = seq(0.5, 9.5, length.out = 8))
df_d <- data.frame(
  x = grid_base$x + rnorm(64, 0, 0.15),
  y = grid_base$y + rnorm(64, 0, 0.15)
)

# Quadrat grid overlay
quadrat_lines <- data.frame(
  x = rep(seq(0, 10, 2.5), each = 2),
  y = rep(c(0, 10), 5),
  group = rep(1:5, each = 2)
)
quadrat_lines_v <- data.frame(
  x = rep(c(0, 10), 5),
  y = rep(seq(0, 10, 2.5), each = 2),
  group = rep(6:10, each = 2)
)

p1 <- ggplot(df_c, aes(x, y)) +
  geom_point(size = 2, color = "red", alpha = 0.6) +
  geom_line(data = quadrat_lines, aes(x, y, group = group), 
            linewidth = 0.8, alpha = 0.5) +
  geom_line(data = quadrat_lines_v, aes(x, y, group = group), 
            linewidth = 0.8, alpha = 0.5) +
  annotate("text", x = 1.25, y = 8.75, label = "15", size = 5, color = "blue") +
  annotate("text", x = 3.75, y = 8.75, label = "2", size = 5, color = "blue") +
  annotate("text", x = 1.25, y = 1.25, label = "14", size = 5, color = "blue") +
  theme_minimal() + ggtitle("Clustered: High Variance") +
  xlim(0, 10) + ylim(0, 10)

p2 <- ggplot(df_r, aes(x, y)) +
  geom_point(size = 2, color = "blue", alpha = 0.6) +
  geom_line(data = quadrat_lines, aes(x, y, group = group), 
            linewidth = 0.8, alpha = 0.5) +
  geom_line(data = quadrat_lines_v, aes(x, y, group = group), 
            linewidth = 0.8, alpha = 0.5) +
  annotate("text", x = 1.25, y = 8.75, label = "4", size = 5, color = "blue") +
  annotate("text", x = 3.75, y = 8.75, label = "5", size = 5, color = "blue") +
  annotate("text", x = 1.25, y = 1.25, label = "3", size = 5, color = "blue") +
  theme_minimal() + ggtitle("Random: Moderate Variance") +
  xlim(0, 10) + ylim(0, 10)

grid.arrange(p1, p2, ncol = 2)
```

**Insight**: Clustered patterns → high variance in cell counts\
Random patterns → moderate variance (Poisson)

------------------------------------------------------------------------

## Variance-Mean Ratio (VMR) {.smaller}

### The Key Statistic

$$VMR = \frac{VAR}{MEAN} = \frac{s^2}{\bar{X}}$$

where: - $s^2$ = variance of cell frequencies - $\bar{X}$ = mean cell frequency

### Theoretical Expectations

| Pattern          | VMR       | Interpretation                            |
|------------------|-----------|-------------------------------------------|
| **Clustered**    | $VMR > 1$ | Variance exceeds mean (overdispersion)    |
| **Random (CSR)** | $VMR = 1$ | Poisson distribution                      |
| **Dispersed**    | $VMR < 1$ | Variance less than mean (underdispersion) |

------------------------------------------------------------------------

## Why VMR = 1 for Random Patterns?

### Poisson Distribution Property

For a **Poisson process**: - Mean = $\lambda$ (events per unit area) - Variance = $\lambda$ - Therefore: $VMR = \lambda / \lambda = 1$

::: callout-important
## Deviation from VMR = 1 indicates departure from CSR

- **VMR \>\> 1**: Strong clustering
- **VMR ≈ 1**: Consistent with random
- **VMR \<\< 1**: Regular/dispersed
:::

------------------------------------------------------------------------

## VMR: Mathematical Details {.smaller}

### Calculating VMR

1.  **Count points in each quadrat**: $f_1, f_2, \ldots, f_m$

2.  **Calculate mean frequency**: $$\bar{X} = \frac{\sum f_i}{m}$$

3.  **Calculate variance**: $$s^2 = \frac{\sum (f_i - \bar{X})^2}{m-1}$$

4.  **Compute VMR**: $$VMR = \frac{s^2}{\bar{X}}$$

------------------------------------------------------------------------

## VMR: Chi-Square Test {.smaller}

### Test Statistic

$$\chi^2 = VMR \times (m - 1)$$

where $m$ = number of quadrats

### Hypothesis Test

- $H_0$: Pattern is random (CSR), $VMR = 1$
- $H_A$: Pattern is not random
  - $H_A: VMR \neq 1$ (two-tailed)
  - $H_A: VMR > 1$ (clustering, one-tailed)
  - $H_A: VMR < 1$ (dispersion, one-tailed)

### Decision Rule

- df = $m - 1$
- Compare $\chi^2$ to critical value from $\chi^2$ distribution
- Common: two-tailed test at $\alpha = 0.05$

------------------------------------------------------------------------

## Quadrat Analysis: Worked Example {.smaller}

**New Mexico Wildfires** (1980-1989)

### Data

- 2,500 random points from 16,000 recorded wildfire starting points
- Study area divided into 120 quadrats (cells)
- Cell frequency mean = 20.9 points/cell
- Variance = 10.5

### Calculations

$$VMR = \frac{10.5}{20.9} = 0.504$$

$$\chi^2 = 0.504 \times (120 - 1) = 0.504 \times 119 = 60.0$$

### Result

- Critical value: $\chi^2_{0.05, 119}$ ≈ 146 (two-tailed)
- Observed: 60.0 \<\< 146
- **Conclusion**: Pattern does NOT differ significantly from random

------------------------------------------------------------------------

## Comparison: Three Decades of Wildfires

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 4

# Simulating wildfire example data from reading
library(knitr)

wildfire_data <- data.frame(
  Decade = c("1980-1989", "1990-1999", "2000-2009"),
  VMR = c(0.504, 20.381, 30.818),
  ChiSquare = c(60.0, 2425, 3667),
  Interpretation = c("Random", "Highly Clustered", "Highly Clustered")
)

kable(wildfire_data, caption = "Wildfire Distribution in New Mexico by Decade")
```

### Insights from Reading

- **1980s**: Random/dispersed pattern
- **1990s & 2000s**: Highly clustered
- **Interpretation**: Human footprint expansion into natural areas
  - Increased clustering over time
  - Evidence of changing fire ignition processes

------------------------------------------------------------------------

## Quadrat Size Selection {.smaller}

::: callout-warning
## Critical Decision

Quadrat size strongly influences results!
:::

### General Guidelines

1.  **Too large**:
    - Many points per cell
    - Low variance → may miss clustering
    - Tends toward VMR = 1
2.  **Too small**:
    - Many empty or single-point cells
    - High variance from sampling variation
    - Unstable estimates
3.  **Rule of thumb**:
    - Mean count per cell: 1-2 points
    - At least $m \geq 50$ quadrats for stable estimates
    - Consider multiple quadrat sizes

------------------------------------------------------------------------

## Quadrat Analysis in R

```{r}
#| echo: true
#| eval: false

# Using spatstat package
library(spatstat)

# Create point pattern
pp <- ppp(x, y, window = owin(c(0, 100), c(0, 100)))

# Quadrat test with specified grid
Q <- quadratcount(pp, nx = 5, ny = 5)  # 5x5 grid = 25 quadrats

# Visual display
plot(Q)
plot(pp, add = TRUE, pch = 16, cex = 0.5)

# Formal test
qt <- quadrat.test(pp, nx = 5, ny = 5)
print(qt)

# Manual VMR calculation
counts <- as.vector(Q)
mean_count <- mean(counts)
var_count <- var(counts)
vmr <- var_count / mean_count
chi_sq <- vmr * (length(counts) - 1)

cat("VMR =", vmr, "\n")
cat("Chi-square =", chi_sq, "\n")
```

------------------------------------------------------------------------

## Comparing NNA and Quadrat Analysis {.smaller}

| Feature             | Nearest Neighbor       | Quadrat                    |
|---------------------|------------------------|----------------------------|
| **Measures**        | Spacing between points | Density variation          |
| **Sensitive to**    | Second-order effects   | First-order effects        |
| **Index**           | R-statistic            | VMR                        |
| **Random baseline** | $R = 1$                | $VMR = 1$                  |
| **Clustered**       | $R < 1$                | $VMR > 1$                  |
| **Dispersed**       | $R > 1$                | $VMR < 1$                  |
| **Advantages**      | Simple, intuitive      | Shows spatial distribution |
| **Limitations**     | Single summary value   | Sensitive to quadrat size  |
| **Best for**        | Overall spacing        | Density variation          |

------------------------------------------------------------------------

## When Results Disagree {.smaller}

### Scenario: NNA says "clustered", Quadrat says "random"

**Possible Explanation**: - Local clustering (detected by NNA) - But overall density relatively uniform across study area - Second-order effects without strong first-order effects

### Scenario: NNA says "random", Quadrat says "clustered"

**Possible Explanation**: - Large-scale density variation (first-order) - But within high-density regions, spacing is random - No local point-to-point interaction

::: callout-tip
**This is why we use multiple methods!** They provide complementary information.
:::

------------------------------------------------------------------------

## Case Study: Toronto Public Services {.smaller}

**From your reading** (Lembo Chapter 14)

### Facilities Analyzed

- Police stations
- Fire stations
- Elementary schools
- Voting locations

### Key Findings

| Facility | NNA Result | Interpretation |
|-------------------|-----------------------|------------------------------|
| **Police** | $R = 1.28$ (dispersed) | Planned spacing for coverage |
| **Fire** | $R = 1.28$ (dispersed) | Optimal response time locations |
| **Elementary Schools** | $R = 0.84$ (clustered) | Follow population density |
| **Voting** | $R = 0.84$ (clustered) | Dense in populated areas |

------------------------------------------------------------------------

## Case Study: Patterns and Processes

### Emergency Services (Police, Fire)

**Pattern**: Dispersed\
**Process**: - Optimization algorithms - Coverage requirements - Response time goals - Planning decisions

### Community Services (Schools, Voting)

**Pattern**: Clustered\
**Process**: - Population distribution - Demand-based location - Accessibility requirements - Historical development

------------------------------------------------------------------------

## Boundary and Study Area Issues {.smaller}

### Critical Considerations

1.  **Study area definition**
    - Arbitrary vs. natural boundaries
    - Affects density calculations
    - Influences null hypothesis
2.  **Edge effects**
    - Points near boundaries have truncated neighborhoods
    - Bias in nearest neighbor distances
    - Solutions: guard areas, corrections
3.  **Scale dependence**
    - Patterns may appear at multiple scales
    - Analysis scale affects results
    - Consider hierarchical approaches

::: callout-important
**Always report**: Study area definition, size, and boundary rationale
:::

------------------------------------------------------------------------

## Practical Workflow for Point Pattern Analysis

### Step-by-Step Approach

1.  **Visualize** the data
    - Create map
    - Visual inspection for obvious patterns
2.  **Calculate density**
    - Points per unit area
    - Kernel density surface
3.  **Run multiple tests**
    - NNA for spacing
    - Quadrat for density variation
    - Consider different quadrat sizes
4.  **Interpret jointly**
    - Do methods agree?
    - First-order vs. second-order?
5.  **Connect to theory**
    - What process could generate this pattern?

------------------------------------------------------------------------

## Advanced Topics (Brief Overview)

### K-function and L-function

- Multi-distance analysis
- Detects clustering at different scales
- More powerful than single-distance NNA

### Marked Point Patterns

- Points have attributes (marks)
- Test for relationships between marks and location
- Example: Tree species and spatial pattern

### Space-Time Point Patterns

- Events have time stamps
- Test for space-time clustering
- Example: Disease outbreak detection

::: notes
These are beyond our scope, but good to know they exist for future work
:::

------------------------------------------------------------------------

## Common Pitfalls and Misconceptions {.smaller}

### ❌ Avoid These Mistakes

1.  **"Random means evenly spread"**
    - No! Random (CSR) shows local clustering by chance
2.  **"Clustering always means interaction"**
    - Could be first-order (density) not process
3.  **"One test is enough"**
    - Use multiple complementary methods
4.  **"P-value tells you the cause"**
    - Significance ≠ explanation
5.  **"Ignore study area"**
    - Boundary definition matters!
6.  **"Assume stationarity"**
    - Check for large-scale trends first

------------------------------------------------------------------------

## R Packages for Point Pattern Analysis

### Primary Package: `spatstat`

```{r}
#| echo: true
#| eval: false

library(spatstat)

# Core functions
ppp()              # Create point pattern
nndist()          # Nearest neighbor distances
quadratcount()    # Quadrat counts
quadrat.test()    # Quadrat test
Kest()            # K-function
density.ppp()     # Kernel density
envelope()        # Simulation envelopes
```

### Other Useful Packages

- `sf`: Spatial data handling
- `mapview`: Interactive mapping
- `splancs`: Spatial and space-time analysis

------------------------------------------------------------------------

# Summary: Lecture 2 {background-color="#660000"}

## Key Concepts Covered

1.  **Quadrat Analysis**
    - Measures density variation
    - Variance-Mean Ratio (VMR)
    - Chi-square test
2.  **Comparing Methods**
    - NNA vs. Quadrat
    - First-order vs. second-order
    - Complementary insights
3.  **Practical Considerations**
    - Boundary effects
    - Scale selection
    - Study area definition
4.  **Case Studies**
    - Toronto public services
    - New Mexico wildfires
    - Interpreting real patterns

------------------------------------------------------------------------

## Before Next Class

### Required Reading

- Review Lembo Chapter 15 (Area Pattern Analysis)
- Focus on Join Count and Moran's I

### Prepare

- Think about areal data in your field
- Examples: Census data, election results, disease rates by county

### Next Lecture

- **Area Pattern Analysis**
- Join Count Analysis (categorical data)
- Moran's I (continuous data)
- Local vs. global spatial autocorrelation

------------------------------------------------------------------------

# Lecture 3: Area Pattern Analysis {background-color="#003366"}

## Transition: Points to Areas {.smaller}

::::: columns
::: {.column width="50%"}
### Point Pattern Data

- Discrete events/objects
- $(x, y)$ coordinates
- Questions:
  - Clustered or dispersed?
  - Random spacing?

### Methods

- Nearest Neighbor
- Quadrat Analysis
:::

::: {.column width="50%"}
### Areal Pattern Data

- **Continuous** spatial units
- Polygons (counties, tracts, regions)
- Questions:
  - Do adjacent areas have similar values?
  - Spatial clustering of attributes?

### Methods (Today)

- **Join Count Analysis**
- **Moran's I** (global & local)
:::
:::::

------------------------------------------------------------------------

## What is Areal Data? {.smaller}

### Definition

**Attributes measured or aggregated over spatial units**

### Characteristics

- **Complete coverage** of study area
- **Mutually exclusive** units
- Values represent entire area (not points)
- **Neighborhood structure** defined by adjacency/proximity

### Examples

| Domain        | Spatial Unit  | Variable        |
|---------------|---------------|-----------------|
| Public Health | Counties      | Disease rate    |
| Economics     | States        | Unemployment    |
| Politics      | Precincts     | Vote share      |
| Environment   | Watersheds    | Pollution level |
| Demography    | Census tracts | Median income   |

------------------------------------------------------------------------

## Key Difference from Point Patterns

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 4

library(ggplot2)
library(gridExtra)

# Point pattern
set.seed(123)
pts <- data.frame(x = runif(40, 0, 10), y = runif(40, 0, 10))

p1 <- ggplot(pts, aes(x, y)) +
  geom_point(size = 3, color = "blue") +
  theme_minimal() +
  ggtitle("Point Pattern") +
  labs(subtitle = "WHERE are the events?")

# Areal data - create simple polygons
poly_data <- data.frame(
  x = c(0,2,2,0,0, 2,4,4,2,2, 4,6,6,4,4, 6,8,8,6,6,
        0,2,2,0,0, 2,4,4,2,2, 4,6,6,4,4, 6,8,8,6,6),
  y = c(0,0,2,2,0, 0,0,2,2,0, 0,0,2,2,0, 0,0,2,2,0,
        2,2,4,4,2, 2,2,4,4,2, 2,2,4,4,2, 2,2,4,4,2),
  group = rep(1:8, each = 5),
  value = rep(c(45, 67, 23, 78, 12, 89, 34, 56), each = 5)
)

p2 <- ggplot(poly_data, aes(x, y, fill = value, group = group)) +
  geom_polygon(color = "black", linewidth = 1) +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  theme_minimal() +
  ggtitle("Areal Data") +
  labs(subtitle = "WHAT are the attribute values?", fill = "Value") +
  coord_fixed()

grid.arrange(p1, p2, ncol = 2)
```

**Point patterns**: Location of events is the variable\
**Areal data**: Attribute values across locations

------------------------------------------------------------------------

## Spatial Autocorrelation {.smaller}

### Core Concept

::: callout-important
## Tobler's First Law of Geography

"Everything is related to everything else, but near things are more related than distant things."
:::

### Spatial Autocorrelation Defined

**Correlation of a variable with itself across space**

- **Positive**: Similar values cluster spatially
  - High near high, low near low
- **Negative**: Dissimilar values are adjacent
  - High near low, low near high
- **Zero**: No spatial pattern (independence)

------------------------------------------------------------------------

## Why Spatial Autocorrelation Matters {.smaller}

### Theoretical Importance

- **Diffusion processes**: Contagion, influence, spread
- **Common exposures**: Environmental factors affect neighbors
- **Spatial spillovers**: Economic/social interactions

### Statistical Implications

- Violates **independence** assumption of classical statistics
- Standard errors underestimated
- Significance tests invalid
- Need spatial statistical methods

### Practical Applications

- Identify spatial clusters (hot spots)
- Disease surveillance
- Environmental justice
- Market analysis

------------------------------------------------------------------------

## Defining Spatial Neighbors {.smaller}

::: callout-note
## Fundamental Question

Who is a neighbor? This defines the spatial structure!
:::

### Common Approaches

1.  **Contiguity** (shared boundary)
    - **Rook**: Shared edge
    - **Queen**: Shared edge or vertex
    - **Higher-order**: Neighbors of neighbors
2.  **Distance**
    - Fixed distance threshold
    - K-nearest neighbors
    - Inverse distance weighting
3.  **Other**
    - Graph-based (networks)
    - Time-based (diffusion)

------------------------------------------------------------------------

## Contiguity Definitions: Visual

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4

# Create simple grid to show rook vs queen
grid <- expand.grid(x = 0:2, y = 0:2)
grid$id <- 1:9

# Central cell is #5
center_cell <- data.frame(x = c(1, 2, 2, 1, 1), y = c(1, 1, 2, 2, 1))

# Rook neighbors (4 neighbors)
rook_neighbors <- data.frame(
  x = c(0,1,1,0,0, 2,3,3,2,2, 1,2,2,1,1, 1,2,2,1,1),
  y = c(1,1,2,2,1, 1,1,2,2,1, 0,0,1,1,0, 2,2,3,3,2),
  group = rep(1:4, each = 5)
)

p1 <- ggplot() +
  geom_rect(xmin = 0, xmax = 3, ymin = 0, ymax = 3, 
            fill = NA, color = "gray50") +
  geom_polygon(data = center_cell, aes(x, y), 
               fill = "yellow", alpha = 0.7, color = "black", linewidth = 1) +
  geom_polygon(data = rook_neighbors, aes(x, y, group = group), 
               fill = "lightblue", alpha = 0.5, color = "black", linewidth = 0.5) +
  geom_vline(xintercept = 0:3, color = "gray30") +
  geom_hline(yintercept = 0:3, color = "gray30") +
  annotate("text", x = 1.5, y = 1.5, label = "A", size = 8) +
  ggtitle("ROOK: Shared Edge (4 neighbors)") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5)) +
  coord_fixed()

# Queen neighbors (8 neighbors)
queen_neighbors <- data.frame(
  x = c(0,1,1,0,0, 2,3,3,2,2, 1,2,2,1,1, 1,2,2,1,1,
        0,1,1,0,0, 2,3,3,2,2, 0,1,1,0,0, 2,3,3,2,2),
  y = c(1,1,2,2,1, 1,1,2,2,1, 0,0,1,1,0, 2,2,3,3,2,
        0,0,1,1,0, 0,0,1,1,0, 2,2,3,3,2, 2,2,3,3,2),
  group = rep(1:8, each = 5)
)

p2 <- ggplot() +
  geom_rect(xmin = 0, xmax = 3, ymin = 0, ymax = 3, 
            fill = NA, color = "gray50") +
  geom_polygon(data = center_cell, aes(x, y), 
               fill = "yellow", alpha = 0.7, color = "black", linewidth = 1) +
  geom_polygon(data = queen_neighbors, aes(x, y, group = group), 
               fill = "lightblue", alpha = 0.5, color = "black", linewidth = 0.5) +
  geom_vline(xintercept = 0:3, color = "gray30") +
  geom_hline(yintercept = 0:3, color = "gray30") +
  annotate("text", x = 1.5, y = 1.5, label = "A", size = 8) +
  ggtitle("QUEEN: Shared Edge or Vertex (8 neighbors)") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5)) +
  coord_fixed()

grid.arrange(p1, p2, ncol = 2)
```

------------------------------------------------------------------------

## Spatial Weights Matrix {.smaller}

### Definition

**W = matrix encoding neighborhood structure**

- Rows and columns = spatial units
- $w_{ij}$ = weight between unit $i$ and unit $j$
- Diagonal = 0 (unit not neighbor to itself)

### Binary Weights (Contiguity)

$$w_{ij} = \begin{cases} 
1 & \text{if } i \text{ and } j \text{ are neighbors} \\
0 & \text{otherwise}
\end{cases}$$

### Row Standardization (Common)

$$w_{ij}^* = \frac{w_{ij}}{\sum_j w_{ij}}$$

Row sums = 1 (weights represent proportions)

------------------------------------------------------------------------

## Example: Simple Weights Matrix

```{r}
#| echo: false

library(knitr)

# Create example for 4 areas
W_example <- data.frame(
  Area = c("A", "B", "C", "D"),
  A = c(0, 1, 1, 0),
  B = c(1, 0, 0, 1),
  C = c(1, 0, 0, 1),
  D = c(0, 1, 1, 0)
)

kable(W_example, caption = "Binary Spatial Weights Matrix (Queen Contiguity)")
```

**Interpretation**: - A is neighbor to B and C - B is neighbor to A and D - C is neighbor to A and D - D is neighbor to B and C

------------------------------------------------------------------------

# Join Count Analysis {background-color="#004d00"}

## Join Count Analysis: Introduction {.smaller}

### Purpose

**Analyze spatial patterns of CATEGORICAL data**

### Key Concept

A **"join"** is a connection between two adjacent areas

### Types of Joins

| Join Type            | Description                              |
|----------------------|------------------------------------------|
| **Black-Black (BB)** | Two adjacent "black" areas               |
| **White-White (WW)** | Two adjacent "white" areas               |
| **Black-White (BW)** | Adjacent areas with different categories |

### The Question

Are there more same-category joins than expected by chance?

------------------------------------------------------------------------

## Join Count: The Logic {.smaller}

### Observed vs. Expected

**Observed**: Count actual joins in the data

**Expected**: What would we expect under random arrangement?

### Hypotheses

- $H_0$: Spatial pattern is random (no spatial autocorrelation)
- $H_A$: Clustering exists
  - **Positive autocorrelation**: More BB and WW joins than expected
  - **Negative autocorrelation**: More BW joins than expected

------------------------------------------------------------------------

## Join Count: Visual Example

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 4

# Create three patterns: clustered, random, dispersed
# Clustered
clust_vals <- c(1,1,1,1,0,0,0,0,0,
                1,1,1,1,0,0,0,0,0,
                1,1,1,1,0,0,0,0,0,
                1,1,1,1,0,0,0,0,0,
                0,0,0,0,0,1,1,1,1)

clust_df <- data.frame(
  x = rep(0:8, times = 5) + 0.5,
  y = rep(0:4, each = 9) + 0.5,
  value = clust_vals
)

# Random
set.seed(999)
rand_df <- data.frame(
  x = rep(0:8, times = 5) + 0.5,
  y = rep(0:4, each = 9) + 0.5,
  value = sample(0:1, 45, replace = TRUE, prob = c(0.5, 0.5))
)

p1 <- ggplot(clust_df, aes(x, y, fill = factor(value))) +
  geom_tile(color = "white", linewidth = 2) +
  scale_fill_manual(values = c("0" = "white", "1" = "black"),
                    labels = c("White", "Black")) +
  theme_void() +
  theme(legend.position = "bottom") +
  ggtitle("Clustered (Positive SA)") +
  labs(subtitle = "Many BB and WW joins", fill = "") +
  coord_fixed()

p2 <- ggplot(rand_df, aes(x, y, fill = factor(value))) +
  geom_tile(color = "white", linewidth = 2) +
  scale_fill_manual(values = c("0" = "white", "1" = "black"),
                    labels = c("White", "Black")) +
  theme_void() +
  theme(legend.position = "bottom") +
  ggtitle("Random (No SA)") +
  labs(subtitle = "Expected number of joins", fill = "") +
  coord_fixed()

grid.arrange(p1, p2, ncol = 2)
```

------------------------------------------------------------------------

## Join Count: Formulas {.smaller}

### Expected Number of Joins (Free Sampling)

For **Black-White joins**:

$$E[BW] = \frac{2JBW}{N(N-1)}$$

where: - $J$ = total number of joins - $B$ = number of black areas - $W$ = number of white areas\
- $N$ = total areas ($B + W$)

### Standard Error

$$\sigma_{BW} = \sqrt{E_{BW} + \frac{\sum L(L-1)BW - 2JBW^2}{N(N-1)} + \frac{4[\sum(L-1)^2] B(B-1)W(W-1)}{N(N-1)(N-2)(N-3)}}$$

where $L$ = number of links (neighbors) for each area

------------------------------------------------------------------------

## Join Count: Test Statistic

### Z-Score

$$Z_{join} = \frac{O_{BW} - E_{BW}}{\sigma_{BW}}$$

where: - $O_{BW}$ = observed number of Black-White joins - $E_{BW}$ = expected number under randomness

### Interpretation

- **Positive** $Z$: More BW joins than expected (negative SA)
- **Negative** $Z$: Fewer BW joins than expected (positive SA / clustering)
- $|Z| > 1.96$ → Significant at $\alpha = 0.05$

------------------------------------------------------------------------

## Join Count Example: US Obesity {.smaller}

**From your reading**: State obesity levels 1995-2020

### Setup

- States classified as **above** or **below** national average
- Queen contiguity weights
- Binary categorization per year

### Results (2020)

| Join Type       | Observed | Expected |
|-----------------|----------|----------|
| **Above-Above** | 26       | 21       |
| **Below-Below** | 27       | 32       |
| **Dissimilar**  | 22       | 53.75    |

**Test**: $Z = -6.40$, $p < 0.001$

**Conclusion**: Significant **positive spatial autocorrelation**\
States with similar obesity levels cluster together

------------------------------------------------------------------------

## Join Count: Interpretation Tips {.smaller}

### What Positive SA Means

- Similar values cluster spatially
- Neighbors tend to be alike
- Evidence of:
  - Diffusion processes
  - Common exposures
  - Spatial spillovers

### What Negative SA Means (Rare)

- Dissimilar values adjacent
- "Checkerboard" pattern
- Possible:
  - Competition effects
  - Complementarity
  - Zoning/planning

### Non-Significance

- Random spatial arrangement
- No evidence of spatial pattern
- Independence assumption reasonable

------------------------------------------------------------------------

## Join Count in R

```{r}
#| echo: true
#| eval: false

# Using spdep package
library(spdep)
library(sf)

# Read spatial data
states <- st_read("states.shp")

# Create binary variable
states$obese_binary <- ifelse(states$obesity > mean(states$obesity), 
                              "High", "Low")

# Create queen contiguity neighbors
nb <- poly2nb(states, queen = TRUE)

# Convert to spatial weights
listw <- nb2listw(nb, style = "B")  # Binary weights

# Join count test
jc_test <- joincount.test(
  fx = factor(states$obese_binary),
  listw = listw
)

print(jc_test)
```

------------------------------------------------------------------------

# Moran's I {background-color="#004d00"}

## Moran's I: Introduction {.smaller}

### Purpose

**Measure spatial autocorrelation for CONTINUOUS data**

### Key Differences from Join Count

- **Join Count**: Categorical (black/white, yes/no)
- **Moran's I**: Continuous (income, temperature, rates)

### Global Moran's I

- **Single value** summarizing spatial autocorrelation for entire study area
- Analogous to Pearson correlation, but for spatial data

### Range

- Typically: $-1 \leq I \leq +1$
- $I > 0$: Positive SA (clustering of similar values)
- $I \approx 0$: No SA (random)
- $I < 0$: Negative SA (dissimilar values adjacent)

------------------------------------------------------------------------

## Moran's I: The Formula {.smaller}

### Global Moran's I

$$I = \frac{n}{\sum_{i}\sum_{j} w_{ij}} \cdot \frac{\sum_{i}\sum_{j} w_{ij}(x_i - \bar{x})(x_j - \bar{x})}{\sum_{i}(x_i - \bar{x})^2}$$

### Simpler form (row-standardized weights):

$$I = \frac{n \sum_{i}\sum_{j} w_{ij}(x_i - \bar{x})(x_j - \bar{x})}{(\sum_{i}\sum_{j} w_{ij}) \sum_{i}(x_i - \bar{x})^2}$$

where: - $n$ = number of areas - $x_i$ = value for area $i$ - $\bar{x}$ = mean value - $w_{ij}$ = spatial weight between $i$ and $j$

------------------------------------------------------------------------

## Moran's I: Intuition {.smaller}

### What It Calculates

**Product of deviations for neighboring areas**

### Positive I (Clustering)

- High values near high values: $(+)(+) = +$
- Low values near low values: $(-)(-)  = +$
- Cross-products tend to be positive

### Negative I (Dispersion)

- High values near low values: $(+)(-) = -$
- Low values near high values: $(-)(+) = -$
- Cross-products tend to be negative

### Random (I ≈ 0)

- Mix of positive and negative cross-products
- Cancel out on average

------------------------------------------------------------------------

## Moran's I: Hypothesis Test {.smaller}

### Expected Value Under Randomness

$$E[I] = \frac{-1}{n-1}$$

(Not zero! Small negative value)

### Variance (Normalization Assumption)

$$Var[I] = \frac{n^2 \cdot \sum_{i} (n - \tilde{n}_i - 1) + n \sum_{i} \tilde{n}_i (n - \tilde{n}_i - 3) - n^2 \sum_{i} \tilde{n}_i^2}{(n-1)^2 \sum_{i} \tilde{n}_i}$$

### Z-Score

$$Z = \frac{I - E[I]}{\sqrt{Var[I]}}$$

### Decision

- $|Z| > 1.96$ → Significant at $\alpha = 0.05$
- Positive $Z$: significant positive autocorrelation
- Negative $Z$: significant negative autocorrelation

------------------------------------------------------------------------

## Moran's I: Worked Example {.smaller}

**Pennsylvania County Obesity Levels** (from reading)

### Setup

- $n = 67$ counties
- Variable: Obesity rate (%)
- Queen contiguity weights

### Results

| Statistic  | Value    |
|------------|----------|
| Moran's I  | 0.69     |
| Expected I | -0.015   |
| Variance   | 0.0043   |
| Z-score    | 10.8     |
| p-value    | \< 0.001 |

**Interpretation**: **Strong positive spatial autocorrelation**\
Counties with similar obesity rates cluster together

------------------------------------------------------------------------

## Global vs. Local Measures {.smaller}

::::: columns
::: {.column width="50%"}
### Global Moran's I

- **One value** for entire study area
- "Is there clustering?"
- Like a correlation coefficient
- Summary statistic

### Limitations

- Doesn't tell you WHERE
- May miss local variation
- Average over whole area
:::

::: {.column width="50%"}
### Local Moran's I (LISA)

- **One value per area**
- "Where is clustering?"
- Identifies hot spots / cold spots
- Spatial outliers

### Advantages

- Maps of local patterns
- Identifies specific locations
- Heterogeneous patterns
:::
:::::

::: callout-tip
Use **global** to test for overall SA, then **local** to find where
:::

------------------------------------------------------------------------

## Local Moran's I (LISA) {.smaller}

### Formula

$$I_i = \frac{(x_i - \bar{x})}{S_i^2} \sum_{j} w_{ij}(x_j - \bar{x})$$

where: - $I_i$ = local Moran's I for area $i$ - $S_i^2 = \frac{\sum (x_i - \bar{x})^2}{n}$ (variance) - $w_{ij}$ = spatial weights

### Interpretation for Area $i$

- **High positive** $I_i$: Area similar to neighbors
  - **High-High**: High value surrounded by high
  - **Low-Low**: Low value surrounded by low
- **Negative** $I_i$: Area different from neighbors
  - **High-Low**: High value surrounded by low (outlier)
  - **Low-High**: Low value surrounded by high (outlier)

------------------------------------------------------------------------

## LISA: Cluster Types

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5

# Create example LISA map
set.seed(123)
n_areas <- 50
lisa_data <- data.frame(
  x = runif(n_areas, 0, 10),
  y = runif(n_areas, 0, 10)
)

# Assign cluster types with spatial structure
lisa_data$cluster_type <- "Not Significant"
# High-High cluster (upper right)
lisa_data$cluster_type[lisa_data$x > 7 & lisa_data$y > 7] <- "High-High"
# Low-Low cluster (lower left)
lisa_data$cluster_type[lisa_data$x < 3 & lisa_data$y < 3] <- "Low-Low"
# Add some outliers
lisa_data$cluster_type[lisa_data$x > 8 & lisa_data$y < 2] <- "High-Low"
lisa_data$cluster_type[lisa_data$x < 2 & lisa_data$y > 8] <- "Low-High"

lisa_colors <- c("High-High" = "red",
                 "Low-Low" = "blue",
                 "High-Low" = "pink",
                 "Low-High" = "lightblue",
                 "Not Significant" = "gray90")

ggplot(lisa_data, aes(x, y, color = cluster_type, size = cluster_type != "Not Significant")) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = lisa_colors) +
  scale_size_manual(values = c("TRUE" = 4, "FALSE" = 2), guide = "none") +
  theme_minimal() +
  labs(title = "LISA Cluster Map",
       subtitle = "Local Indicators of Spatial Association",
       color = "Cluster Type",
       x = "", y = "") +
  theme(legend.position = "bottom")
```

------------------------------------------------------------------------

## LISA: Cleveland Racial Composition {.smaller}

**Example from reading**: Census block groups

### Four Groups Analyzed

- Asian, Black, White, Hispanic

### Key Findings (Moran's I values)

| Group        | Moran's I | Z     | p-value | Interpretation         |
|--------------|-----------|-------|---------|------------------------|
| **Asian**    | 0.35      | 18.39 | 0.000   | Moderate clustering    |
| **Black**    | 0.58      | 64.6  | 0.000   | Strong clustering      |
| **White**    | 0.64      | 65.4  | 0.000   | Strong clustering      |
| **Hispanic** | 0.81      | 65.7  | 0.000   | Very strong clustering |

**Conclusion**: Significant residential segregation across all groups\
Hispanic communities most strongly clustered

------------------------------------------------------------------------

## Moran Scatterplot {.smaller}

### Visual Tool for Understanding Moran's I

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6

# Create example Moran scatterplot
set.seed(456)
n <- 100
x <- rnorm(n)
# Create spatially lagged y with positive autocorrelation
y <- 0.7 * x + rnorm(n, 0, 0.5)

moran_df <- data.frame(
  value = x,
  lagged_value = y
)

# Assign quadrants
moran_df$quadrant <- "HH"
moran_df$quadrant[moran_df$value < 0 & moran_df$lagged_value > 0] <- "LH"
moran_df$quadrant[moran_df$value < 0 & moran_df$lagged_value < 0] <- "LL"
moran_df$quadrant[moran_df$value > 0 & moran_df$lagged_value < 0] <- "HL"

ggplot(moran_df, aes(value, lagged_value, color = quadrant)) +
  geom_point(size = 2, alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("HH" = "red", "LL" = "blue", 
                                 "HL" = "pink", "LH" = "lightblue")) +
  theme_minimal() +
  labs(title = "Moran Scatterplot",
       subtitle = "Slope = Moran's I",
       x = "Value (Standardized)",
       y = "Spatially Lagged Value",
       color = "Quadrant") +
  annotate("text", x = 2, y = 2, label = "High-High", size = 5) +
  annotate("text", x = -2, y = -2, label = "Low-Low", size = 5) +
  annotate("text", x = 2, y = -2, label = "High-Low", size = 5) +
  annotate("text", x = -2, y = 2, label = "Low-High", size = 5)
```

**Slope of regression line = Moran's I**

------------------------------------------------------------------------

## Moran's I in R {.smaller}

```{r}
#| echo: true
#| eval: false

library(spdep)
library(sf)

# Read spatial data
counties <- st_read("counties.shp")

# Create neighbors list (Queen contiguity)
nb <- poly2nb(counties, queen = TRUE)

# Create spatial weights (row-standardized)
listw <- nb2listw(nb, style = "W")

# Global Moran's I test
moran_test <- moran.test(counties$obesity, listw)
print(moran_test)

# Moran scatterplot
moran.plot(counties$obesity, listw, 
           main = "Moran Scatterplot: Obesity",
           xlab = "Obesity Rate (standardized)",
           ylab = "Spatially Lagged Obesity")

# Local Moran's I (LISA)
lisa <- localmoran(counties$obesity, listw)
counties$lisa_I <- lisa[, 1]
counties$lisa_p <- lisa[, 5]

# Map significant clusters
plot(counties["lisa_I"], main = "Local Moran's I")
```

------------------------------------------------------------------------

## LISA Significance and Multiple Testing {.smaller}

::: callout-warning
## Problem: Multiple Comparisons

Testing hundreds of areas → inflation of Type I error!
:::

### Solutions

1.  **Bonferroni correction**
    - $\alpha_{adj} = \alpha / n$
    - Very conservative
2.  **False Discovery Rate (FDR)**
    - Controls proportion of false positives
    - Less conservative than Bonferroni
3.  **Permutation approach**
    - Simulate null distribution
    - More robust

### Recommendation

- Report both corrected and uncorrected p-values
- Use permutation tests (available in `spdep`)

------------------------------------------------------------------------

## Comparing Area Pattern Methods {.smaller}

| Feature            | Join Count                | Moran's I                 |
|--------------------|---------------------------|---------------------------|
| **Data type**      | Categorical               | Continuous                |
| **Output**         | Join counts               | Correlation-like index    |
| **Range**          | Count values              | -1 to +1                  |
| **Interpretation** | Number of joins           | Degree of autocorrelation |
| **Local version**  | No standard version       | LISA (Local Moran's I)    |
| **Use cases**      | Land use, binary outcomes | Rates, prices, densities  |

### Complementary Approaches

- Both test spatial autocorrelation
- Both require defining neighbors
- Both compare observed to expected under randomness

------------------------------------------------------------------------

## Practical Applications {.smaller}

### Public Health

- **Disease mapping**: Identify disease clusters
- **Health disparities**: Unequal access to care
- **Outbreak detection**: Early warning systems

### Urban Planning

- **Gentrification**: Spreading neighborhood change
- **Segregation**: Residential clustering by race/income
- **Housing markets**: Price spillovers

### Environmental

- **Pollution**: Clustering of environmental hazards
- **Environmental justice**: Disproportionate exposure
- **Climate adaptation**: Regional vulnerability

### Economics

- **Unemployment**: Regional labor markets
- **Innovation**: Technology spillovers
- **Poverty**: Persistent spatial poverty traps

------------------------------------------------------------------------

## Common Issues and Solutions {.smaller}

### Issue 1: Modifiable Areal Unit Problem (MAUP)

- Results sensitive to zone definition
- **Solution**: Test multiple aggregations, report sensitivity

### Issue 2: Boundary Definition

- Edge areas have fewer neighbors
- **Solution**: Consider broader study area, report boundary effects

### Issue 3: Choosing Neighbors

- Contiguity vs. distance? Rook vs. Queen?
- **Solution**: Test multiple definitions, justify choice theoretically

### Issue 4: Statistical vs. Practical Significance

- Large samples → everything significant
- **Solution**: Report effect sizes (Moran's I value), visualize

------------------------------------------------------------------------

## Assumptions and Limitations {.smaller}

### Key Assumptions

1.  **Stationarity**: Process constant across space
    - May not hold in heterogeneous regions
2.  **Isotropy**: Relationships same in all directions
    - Geographic barriers may violate this
3.  **Appropriate neighborhood definition**
    - Neighbors must be meaningful
4.  **Independence of areas** (for inference)
    - Actually violated if SA exists! (parametric bootstrap helps)

### Limitations

- **Descriptive, not causal**
- **Cannot distinguish processes** (multiple processes → same pattern)
- **Scale-dependent** (MAUP)

------------------------------------------------------------------------

## Advanced Topics (Brief Overview) {.smaller}

### Other Spatial Autocorrelation Measures

- **Geary's C**: More sensitive to local differences
- **Getis-Ord G**: Hot spot analysis
- **Spatial correlogram**: SA at multiple distances

### Spatial Regression Models

- When SA exists, OLS invalid
- **Spatial lag model**: Y depends on neighbor Y
- **Spatial error model**: Errors spatially correlated

### Space-Time Analysis

- Extend to temporal dimension
- Emerging/declining clusters
- **Knox test**, **Space-time scan statistics**

::: notes
These go beyond ESDA into confirmatory spatial analysis
:::

------------------------------------------------------------------------

## Best Practices Summary {.smaller}

### Workflow for Area Pattern Analysis

1.  **Visualize**: Create choropleth maps
2.  **Define neighbors**: Justify choice
3.  **Test globally**: Join Count or Moran's I
4.  **If significant, test locally**: LISA
5.  **Map results**: Show cluster locations
6.  **Interpret substantively**: Connect to theory/process

### Reporting Checklist

✓ Study area definition and rationale\
✓ Neighborhood definition (rook/queen/distance)\
✓ Weighting scheme (binary/row-standardized)\
✓ Global test results (statistic, p-value)\
✓ Local analysis (if applicable)\
✓ Maps of patterns\
✓ Substantive interpretation

------------------------------------------------------------------------

## Integration: Point and Area Patterns {.smaller}

### Conceptual Connections

| Concept | Point Pattern | Area Pattern |
|------------------|---------------------------|---------------------------|
| **Random baseline** | CSR | Spatial independence |
| **Clustering** | NNA: R \< 1<br>Quadrat: VMR \> 1 | Positive Moran's I<br>High join counts |
| **Dispersion** | NNA: R \> 1<br>Quadrat: VMR \< 1 | Negative Moran's I |
| **Local vs. Global** | Individual distances vs. overall R | LISA vs. global I |
| **First-order** | Quadrat analysis | Less emphasized |
| **Second-order** | NNA | Moran's I, Join Count |

------------------------------------------------------------------------

# Summary: Lecture 3 {background-color="#660000"}

## Key Concepts Covered {.smaller}

1.  **Areal Data**
    - Continuous spatial units
    - Attribute values over areas
2.  **Spatial Autocorrelation**
    - Tobler's Law
    - Positive, negative, none
3.  **Spatial Weights**
    - Defining neighbors
    - Rook vs. Queen contiguity
4.  **Join Count Analysis**
    - Categorical data
    - Black-white joins
    - Testing for clustering
5.  **Moran's I**
    - Continuous data
    - Global and Local (LISA)
    - Hot spot identification

------------------------------------------------------------------------

## Module Synthesis {.smaller}

### Three Lectures, Unified Framework

**Lecture 1**: Foundations - Patterns vs. processes - First vs. second-order effects - Point patterns: NNA

**Lecture 2**: Point patterns continued - Quadrat analysis - Comparing methods - Applications

**Lecture 3**: Area patterns - Join Count (categorical) - Moran's I (continuous) - Local spatial analysis

### Core Theme

**Exploratory spatial data analysis reveals structure**\
→ Inform modeling and theory development

------------------------------------------------------------------------

## Moving Forward {.smaller}

### From ESDA to Confirmatory Analysis

**ESDA (what we covered)**: - Describe patterns - Generate hypotheses - Identify where to look

**Next steps** (beyond this module): - Spatial regression models - Hypothesis testing with spatial data - Causal inference in spatial context

### Skills You've Gained

✓ Recognize spatial patterns\
✓ Choose appropriate methods\
✓ Interpret spatial statistics\
✓ Connect patterns to processes\
✓ Implement in R\
✓ Report findings effectively

------------------------------------------------------------------------

## Final Recommendations {.smaller}

### For Your Research

1.  **Start with visualization**
    - Maps reveal patterns statistics might miss
2.  **Use multiple methods**
    - Different methods = different insights
3.  **Think about process**
    - What mechanism generated this pattern?
4.  **Consider scale**
    - Patterns are scale-dependent (MAUP)
5.  **Report thoroughly**
    - Study area, neighbors, weights, sensitivity
6.  **Be humble about inference**
    - Pattern ≠ process
    - Correlation ≠ causation
    - Spatial patterns suggest but don't prove

------------------------------------------------------------------------

## Resources for Further Learning {.smaller}

### Textbooks

- Bivand et al. (2013) *Applied Spatial Data Analysis with R*
- Anselin (1988) *Spatial Econometrics*
- O'Sullivan & Unwin (2010) *Geographic Information Analysis*

### Software

- **R packages**: `spatstat`, `spdep`, `sf`, `sp`
- **Python**: `PySAL`, `GeoPandas`
- **GeoDa**: Free software by Luc Anselin

### Online Resources

- CRAN Task View: Spatial
- Geocomputation with R: https://geocompr.robinlovelace.net/
- Spatial Data Science: https://r-spatial.org/

------------------------------------------------------------------------

## Thank You! {.center}

### Questions?

**Office Hours**: \[Your schedule\]\
**Email**: \[Your email\]

### Assignment Due

\[Date and details\]

Good luck with your spatial analysis projects!

------------------------------------------------------------------------

## Appendix: R Code Examples {.smaller visibility="uncounted"}

### Complete Workflow Example

```{r}
#| echo: true
#| eval: false

# Load packages
library(sf)
library(spatstat)
library(spdep)
library(ggplot2)

# === Point Pattern Analysis ===

# Load point data
coords <- read.csv("points.csv")
pp <- ppp(coords$x, coords$y, window = owin(c(0,100), c(0,100)))

# Nearest Neighbor Analysis
nnd_mean <- mean(nndist(pp))
nnd_expected <- 1 / (2 * sqrt(npoints(pp) / area(Window(pp))))
R_stat <- nnd_mean / nnd_expected

# Quadrat Analysis
Q <- quadratcount(pp, nx = 5, ny = 5)
quadrat.test(pp, nx = 5, ny = 5)

# === Area Pattern Analysis ===

# Load areal data
regions <- st_read("regions.shp")

# Create spatial weights
nb <- poly2nb(regions, queen = TRUE)
listw <- nb2listw(nb, style = "W")

# Global Moran's I
moran.test(regions$value, listw)

# Local Moran's I (LISA)
lisa <- localmoran(regions$value, listw)
regions$lisa_I <- lisa[,1]
regions$lisa_p <- lisa[,5]

# Map results
ggplot(regions) +
  geom_sf(aes(fill = lisa_I)) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  theme_minimal()
```
