---
title: "Geographic Data Visualization"
author: "Varun Goel"
date: "`r Sys.Date()`"
format: html
editor: visual
---

# load data and packages
```{r}
library(tidyverse)
library(readr)
library(infer)

dat_2prop <- read_rds("./data/dat_2prop.rds")
```


## Example 2: Does place matter? 

Are urban residents more likely to believe that climate change is caused by humans, as compared to rural residents?
    
Q) **What is our testable hypothesis? Type your Null and Alternate Hypothesis**


## Run the following Desriptive Analysis

```{r}
ggplot(dat_2prop, aes(x = area_type, fill = climhuman)) +
  geom_bar(position = "fill") +
  labs(x = "Area Type", y = "Proportion (Climate change caused by humans)") +
  coord_flip()
```

Based on the graph, what is your guess? Write in 1 sentence

## Calculate Difference in Proportion

This analysis is akin to running hypothesis testing for 2 nominal variables.

```{r}

dat_2prop |>
  group_by(area_type) |>
  summarize(prop_clim = sum(climhuman=="yes")/length(climhuman))

d_hat <- dat_2prop |>
  specify(climhuman ~ area_type, success = "yes") |>
  calculate(stat = "diff in props", order = c("urban", "rural"))
d_hat
```


## Calculate Null Distribution

```{r}
set.seed(2018) 

null_distn_two_props <- dat_2prop |>
  specify(climhuman ~ area_type, success = "yes") |>
  hypothesize(null = "independence") |>
  generate(reps = 10000) |>
  calculate(stat = "diff in props", order = c("urban", "rural"))
```

## Calculate P value

```{r}
null_distn_two_props |>
  visualize() +
  shade_p_value(obs_stat = d_hat, direction = "greater")

pvalue <- null_distn_two_props  |>
  get_pvalue(obs_stat = d_hat, direction = "greater")
pvalue
```

## Calculate Confidence Interval

```{r}
boot_distn_two_props <- dat_2prop |>
  specify(climhuman ~ area_type, success = "yes") |>
  generate(reps = 10000) |>
  calculate(stat = "diff in props", order = c("urban", "rural"))

ci <- boot_distn_two_props |>
  get_ci()
ci
```

## Draw Confidence Interval

```{r}
boot_distn_two_props |>
  visualize() +
  shade_ci(endpoints = ci) +
  geom_vline(xintercept = 0,
             color = "black")
  
```

## If you were to use a statistical formula


```{r echo=TRUE}
prop.test(x = c(45, 24), n = c(60, 40),
              alternative = "greater")
```


## Conclusion

#### Answer the following questions:

1) Do we reject the null hypothesis? Why or why not?
2) Are our results similar to the statistical test?

